---
title: "Session 3: Weather API — Setup, Current Weather & Geocoding"
subtitle: "Day 2 — 9:00 AM to 10:00 AM"
author: "Immanuel Williams"
date: "August 6, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    code-tools: true
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## Welcome Back — Day 2

Yesterday we built the conceptual and practical foundation:

- **Session 1:** JSON structure, request/response flow, status codes
- **Session 2:** CRUD framework, GET/POST methods, URL anatomy, dummy server practice

Today we apply everything to a **real API** — the [OpenWeather API](https://openweathermap.org/api). We'll make authentic requests, handle real responses, and build reusable functions.


---


## Part 1: Setting Up Your API Key

Before we can make requests to the OpenWeather API, we need an **API key**. This is like your library card — it identifies who is making the request.

::: {.callout-important}
## Security Best Practice
**Never** write your API key directly in your R script. If you share that script or push it to GitHub, your key is exposed. Instead, store it in an environment file.
:::

### Step 1: Store Your Key

We'll use `dotenv` to load our API key from a `.Renviron.txt` file:

```{r}
# Create or edit your .Renviron.txt file
# It should contain a single line like:
# API_KEY=your_actual_api_key_here
```

### Step 2: Load and Access the Key

```{r}
dotenv::load_dot_env(file = ".Renviron.txt")

# Verify it loaded (should print your key in the console)
# Sys.getenv("API_KEY")
```


#### ———————
### Q1. {.unnumbered}

**(Discussion):** Why is it a critical security best practice to store your API key in an environment file (`.Renviron`) instead of writing it directly in your R script? What's the main risk of ignoring this practice?

#### ———————


### Step 3: Load Libraries

```{r}
#| output: false
library(httr2)       # Makes web requests
library(tibble)      # Tidyverse version of data.frame
library(lubridate)   # Time and date handling
library(ggplot2)     # Visualizations
library(dplyr)       # Data manipulation
library(dotenv)      # Load environment variables
library(glue)        # Attach strings together
```


---


## Part 2: OpenWeather API Overview

OpenWeather offers several APIs. Here are the three we'll work with:

| Feature | **Current Weather API** | **Geocoding API** | **One Call API** |
|------------------|------------------|------------------|----------------------|
| **Purpose** | Get current weather for a city | Convert city names to coordinates | Get full weather data for a coordinate |
| **Input** | City name, city ID, coordinates, or zip | City name or zip code | Latitude & Longitude |
| **Output** | Temperature, conditions, wind, etc. | Location info (lat/lon, country) | Current, hourly, daily, alerts |
| **Endpoint** | `/data/2.5/weather` | `/geo/1.0/direct` | `/data/3.0/onecall` |


#### ———————
### Q2. {.unnumbered}

**(Multiple Choice — Callback to Session 2 Q6):** Looking at the table above, the **endpoint** for the Current Weather API is `/data/2.5/weather`. If the base URL is `https://api.openweathermap.org`, and you want weather for San Luis Obispo in imperial units, which of the following is the correct full URL structure?

a) `https://api.openweathermap.org?q=San+Luis+Obispo&units=imperial`
b) `https://api.openweathermap.org/data/2.5/weather?q=San+Luis+Obispo&appid=YOUR_KEY&units=imperial`
c) `https://api.openweathermap.org/data/2.5/weather/San+Luis+Obispo`
d) `https://api.openweathermap.org/weather?city=San+Luis+Obispo`

#### ———————


---


## Part 3: Get Current Weather for a City

Let's walk through the complete process of making a real API request.

### Step 1: Select a City

```{r}
city_name <- "San Luis Obispo"
```

### Step 2: Build the Request URL

We use `glue()` to construct the URL by combining the base endpoint with our query parameters:

```{r}
current_weather_url <- glue(
  "https://api.openweathermap.org/data/2.5/weather?",
  "q=", URLencode(city_name),
  "&appid=", Sys.getenv("API_KEY"),
  "&units=", "imperial"
)
```

::: {.callout-note}
## `URLencode()` — Why?
City names with spaces (like "San Luis Obispo") need to be encoded for URLs. `URLencode()` converts spaces to `+` or `%20` so the URL is valid. Without this, the server won't understand your request.
:::


### Step 3: Send the Request

```{r}
req <- request(current_weather_url)
response <- req_perform(req)

# Check what kind of content we got back
resp_content_type(response)
```

The content type should be `application/json` — confirming we received JSON data.


### Step 4: Parse the Response

```{r}
if (resp_status(response) == 200) {
  # Parse JSON into an R list
  result <- resp_body_json(response)
  
  # Convert to a data frame
  currweather_df <- as.data.frame(result)
  
  # View selected columns
  print(select(currweather_df, name, coord.lon, coord.lat, weather.main, main.temp))
} else {
  cat("Failed. Status code:", resp_status(response), "\n")
}
```

::: {.callout-note}
## What Just Happened?
1. We checked the status code — looking for `200 OK`
2. We parsed the JSON response into an R named list using `resp_body_json()`
3. We flattened the nested list into a data frame using `as.data.frame()`
4. The nested JSON keys became column names like `coord.lon`, `weather.main`, `main.temp`
:::


#### ———————
### Q3. {.unnumbered}

**(Open-Ended):** In Step 4, we used `as.data.frame(result)` to convert the JSON. The column `main.temp` comes from a nested JSON structure. What did the original JSON probably look like for that value?

#### ———————


---


## Part 4: Two Ways to Build Request URLs

There's an alternative way to build the request using `httr2`'s query method:

```{r}
# Method 1: Glue (what we used above)
url_glue <- glue(
  "https://api.openweathermap.org/data/2.5/weather?",
  "q=", URLencode(city_name),
  "&appid=", Sys.getenv("API_KEY"),
  "&units=", "imperial"
)

# Method 2: httr2 query parameters
req_query <- request("https://api.openweathermap.org/data/2.5/weather") |>
  req_url_query(
    q = city_name,
    appid = Sys.getenv("API_KEY"),
    units = "imperial"
  )
```

Both produce the same request. The `glue` method helps you **see** the full URL and understand its structure. The `req_url_query` method is cleaner for complex queries.


#### ———————
### Q4. {.unnumbered}

**(Multiple Choice):** Which of the following is an advantage of using the `glue()` method to build URLs, especially when learning about APIs?

a) It automatically handles authentication
b) It makes the full URL structure visible, helping you understand the anatomy of the request
c) It's faster than the query method
d) It automatically encodes special characters

#### ———————


---


## Part 5: Geocoding API — Converting City Names to Coordinates

Some APIs require **latitude and longitude** instead of city names. The OpenWeather Geocoding API converts city names to coordinates.

### Building the Geocode Function

```{r}
geocode <- function(city) {
  
  # Step 1: Build the geocoding URL
  geo_url <- glue(
    "http://api.openweathermap.org/geo/1.0/direct?",
    "q=", URLencode(city),
    "&limit=1&appid=", Sys.getenv("API_KEY")
  )
  
  # Step 2: Make the request
  geo_response <- req_perform(request(geo_url))
  
  # Step 3: Parse if successful
  if (resp_status(geo_response) == 200) {
    geo_data_df <- resp_body_json(geo_response) |>
      as.data.frame()
    
    # Step 4: Check if we got results
    if (length(geo_data_df) == 0) {
      stop("City not found. Please check the city name.")
    }
    
    # Step 5: Extract and round coordinates
    lat <- round(geo_data_df$lat, digits = 2)
    lon <- round(geo_data_df$lon, digits = 2)
    
    # Step 6: Return the coordinates
    cat("Coordinates for", city, "-> Latitude:", lat, "Longitude:", lon, "\n")
    return(list(lat = lat, lon = lon))
  } else {
    stop(paste("Geocoding failed. Status:", resp_status(geo_response)))
  }
}
```


### Test the Function

```{r}
# Test with a known city
geocode("San Luis Obispo, CA, USA")
```

```{r}
# Try your hometown!
geocode("Ames, IA, USA")
```


#### ———————
### Q5. {.unnumbered}

**(Open-Ended — Callback to Session 1 Q7):** In the `geocode` function, we check `if (resp_status(geo_response) == 200)` before parsing. What would happen if we skipped this check and the API returned a `404` status? How does this connect to what we learned about status codes?

#### ———————


#### ———————
### Q6. {.unnumbered}

**(Discussion):** Look at the `geocode` function. It uses `stop()` when the city is not found. What is the difference between `stop()`, `warning()`, and `cat()` for communicating errors in R? When would you use each?

#### ———————


---


## Part 6: Your Turn — Build Your Own Request

Now it's your turn! Using what we've learned, make a current weather request for **your hometown** (or any city you choose).

### Activity: Fill in the Blanks

```{r}
# Step 1: Choose your city
my_city <- "___________"

# Step 2: Build the URL (fill in the missing parts)
my_url <- glue(
  "https://api.openweathermap.org/data/2.5/weather?",
  "q=", URLencode(________),
  "&appid=", Sys.getenv("________"),
  "&units=", "________"
)

# Step 3: Make the request
my_req <- request(my_url)
my_response <- req_perform(my_req)

# Step 4: Check status and parse
if (resp_status(my_response) == ________) {
  my_result <- resp_body_json(my_response)
  my_df <- as.data.frame(my_result)
  
  # Step 5: Print relevant columns
  print(select(my_df, name, main.temp, weather.main, main.humidity))
} else {
  cat("Request failed with status:", resp_status(my_response), "\n")
}
```



---


## Session 3 Recap

In this session, we made the leap to **real API requests**:

1. **API key setup** — storing keys securely in `.Renviron`
2. **Current Weather API** — building URLs, sending requests, parsing JSON responses
3. **Two URL construction methods** — `glue()` vs. `req_url_query()`
4. **Geocoding API** — converting city names to latitude/longitude coordinates
5. **Function building** — encapsulating API logic in reusable functions
6. **Hands-on activity** — making your own weather request

**Up next in Session 4:** We'll build on the geocoding function to retrieve **historical weather data**, loop over multiple days, create visualizations, and wrap up with a reflection!
