---
title: "Session 3: Weather API — First Real Requests & Geocoding"
subtitle: "Day 2 — 9:00 AM to 10:00 AM"
author: "Immanuel Williams"
date: "August 6, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    code-tools: true
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## Welcome Back — Day 2

Yesterday we built the conceptual foundation:

- **Session 1:** JSON structure, request/response flow, status codes
- **Session 2:** CRUD framework, GET/POST methods, URL anatomy

Today we apply everything to a **real API** — the free [Open-Meteo weather API](https://open-meteo.com/). We'll make authentic requests, handle real responses, and turn them into data frames we can analyze.

::: {.callout-important}
## 🎯 The One Goal — Today We Do It

> **Make a request to an API, then format the response into something we can analyze.**

By the end of this hour, you will have done this — for real, in R, with live weather data.
:::


---


## Part 1: Meet the Open-Meteo API

We will use **Open-Meteo** — a free weather API built for exactly this kind of learning:

- **No sign-up. No API key. No credit card.** You can start making requests immediately.
- The main endpoint is:

```{r}
request("https://api.open-meteo.com/v1/forecast")
```

That address — `https://api.open-meteo.com/v1/forecast` — is the **endpoint** from Session 2: the stable "front door" that never changes. Everything else we want (which city, which measurements, which units) goes in the **query string**.

::: {.callout-note}
## What About API Keys, Clients & Secrets?

Many commercial APIs (OpenWeather, X, Spotify, ...) require you to register and receive credentials — an **API key**, sometimes a **client ID** and **secret** — which you attach to every request like an ID card, and which you must store carefully (never directly in your script!). It's an important real-world concept, and the Practice Labs on this site show it in action.

**But today we get to skip all of it.** Open-Meteo is free and open — so we can spend our hour on the actual skill: requests and responses.
:::

::: {.callout-tip}
## 🎯 Goal Check
One endpoint, no key, live data. The only thing between us and the goal is building the query string — which you already learned to read yesterday.
:::


#### ———————
### Q1. {.unnumbered}

**(Fill in the Blank):** The stable part of the URL — `https://api.open-meteo.com/v1/forecast` — that never changes from one request to the next is called the \_\_\_\_\_\_\_\_.

#### ———————


#### ———————
### Q2. {.unnumbered}

**(Multiple Choice):** Which statement about the Open-Meteo API is true?

a) It requires a paid subscription.
b) It requires an API key stored in `.Renviron`.
c) It is free and requires no API key at all.
d) It only works inside Positron.
e) It only returns data for cities in Europe.

#### ———————


---


## Part 2: See It in the Browser First

Here is the payoff of Session 2's URL anatomy: **a request is just a URL — so a web browser can make one.**

Copy this URL and paste it into a new browser tab, exactly as it is:

```
https://api.open-meteo.com/v1/forecast?latitude=35.28&longitude=-120.66&current=temperature_2m&temperature_unit=fahrenheit
```

**Try it now.** What comes back is not a web page — it is raw **JSON**, right there in your browser:

``` json
{
  "latitude": 35.28,
  "longitude": -120.66,
  "timezone": "GMT",
  "current_units": { "time": "iso8601", "temperature_2m": "°F" },
  "current": {
    "time": "2026-08-06T16:00",
    "temperature_2m": 71.4
  }
}
```

Look at what you can already recognize from Sessions 1 and 2:

- The **endpoint** and the **query string** in the address bar (`?`, `&`, key-value pairs)
- The response wrapped in curly braces `{ }` — the **body**
- **Data** (`temperature_2m: 71.4`), and **metadata** (`current_units`, `timezone`)
- Your browser made a **GET** request — typing a URL into the address bar is always a GET

Now change `latitude=35.28&longitude=-120.66` to another location (Boston is `latitude=42.36&longitude=-71.06`) and reload. Live data, different city, same endpoint.

::: {.callout-tip}
## 🎯 Goal Check
You just completed the first half of the goal — **you made a request** — with zero code. R is only going to automate what the browser just did.
:::


#### ———————
### Q3. {.unnumbered}

**(Fill in the Blank):** When you paste the request URL into a browser, what comes back is not a web page — it is raw \_\_\_\_\_\_\_\_.

#### ———————


#### ———————
### Q4. {.unnumbered}

**(Multiple Choice):** Typing a URL into the browser's address bar and pressing Enter performs which kind of HTTP request?

a) POST
b) DELETE
c) PUT
d) PATCH
e) GET

#### ———————


---


## Part 3: The ETV Cycle — Our Map for the Rest of the Workshop

<!-- TODO(Immanuel): link/embed your ETV Cycle 4 notes here -->

Everything we do from here on follows one repeating cycle — **ETV**:

| Step | Name | What it means | In code |
|------|------|----------------|---------|
| **E** | **Extract** | Make the request, receive the JSON response | `request()` → `req_url_query()` → `req_perform()` |
| **T** | **Transform** | Format the response into a data frame | `resp_body_json()` → `tibble()` |
| **V** | **Visualize** | Turn the data frame into a plot | `ggplot()` |

We'll do **E** and **T** in this session and add **V** in Session 4. Whenever you feel lost in code, ask: *which ETV step am I on right now?*

### Load Libraries

```{r}
#| output: false
library(httr2)       # E: makes web requests
library(tibble)      # T: tidyverse data frames
library(dplyr)       # T: data manipulation
library(purrr)       # T: map() — apply a recipe to many inputs
library(ggplot2)     # V: visualizations
```

::: {.callout-tip}
## 🎯 Goal Check
ETV *is* the goal sentence, written as three letters: **E** = make a request, **T** = format the response, **V** = analyze it.
:::


---


## Part 4: Extract — The Browser Request, Now in R

Let's have R do exactly what the browser just did. Two steps to build the request, one to send it:

```{r}
# E is for Extract ------------------------------------------

# Step 1: Start from the endpoint and add the query
req <- request("https://api.open-meteo.com/v1/forecast") |>
  req_url_query(
    latitude = 35.28,
    longitude = -120.66,
    current = "temperature_2m",
    temperature_unit = "fahrenheit"
  )

# Step 2: Look at the URL R built for us — same as the browser one!
req

# Step 3: Send the request
response <- req_perform(req)

# Step 4: Check the status code
resp_status(response)
```

The status should be `200` — success, just like Session 1 taught us to expect.

::: {.callout-note}
## `req_url_query()` Writes the URL for You
Each argument becomes one `key=value` pair in the query string, and `httr2` handles the `?`, the `&`, and any special-character encoding automatically. Printing `req` shows the finished URL — compare it character-by-character with the one you pasted into your browser.
:::

```{r}
# Step 5: Parse the JSON body into a named list
result <- resp_body_json(response)

# Step 6: Peek at the structure
str(result, max.level = 2)
```


#### ———————
### Q5. {.unnumbered}

**(Fill in the Blank):** After `req_perform()`, we call `resp_status(response)` and hope to see the number \_\_\_\_\_\_\_\_.

#### ———————


#### ———————
### Q6. {.unnumbered}

**(Multiple Choice):** What does `req_url_query()` do?

a) It sends the request to the server.
b) It parses the JSON response.
c) It checks the status code.
d) It adds `key=value` pairs to the URL's query string, handling `?`, `&`, and encoding for you.
e) It draws a plot of the response.

#### ———————


---


## Part 5: Transform — From JSON to a Data Frame

The parsed `result` is a **named list** (Session 1!). The piece we care about is `result$current`:

```{r}
# T is for Transform ----------------------------------------

result$current

# Pull the values we want into a tibble
current_df <- tibble(
  city   = "San Luis Obispo",
  time   = result$current$time,
  temp_f = result$current$temperature_2m
)

current_df
```

That's **E** and **T** — the entire workshop goal — in about ten lines.

### Scaling Up: Many Cities with `map()`

What if we want **several cities**? We don't need to copy-paste the code three times, and we don't need to write our own function either. We give `pmap()` a *recipe* — an **anonymous function**, written `\(...)` — and it applies the recipe to every row of a small table:

```{r}
# A small table of cities and their coordinates
cities <- tribble(
  ~city,             ~latitude, ~longitude,
  "San Luis Obispo",     35.28,    -120.66,
  "Boston",              42.36,     -71.06,
  "Ames",                42.03,     -93.62
)

# The recipe: for one row (one city), do E and T, return a one-row tibble
current_df <- cities |>
  pmap(\(city, latitude, longitude) {

    # E: extract
    result <- request("https://api.open-meteo.com/v1/forecast") |>
      req_url_query(
        latitude = latitude,
        longitude = longitude,
        current = "temperature_2m",
        temperature_unit = "fahrenheit"
      ) |>
      req_perform() |>
      resp_body_json()

    # T: transform
    tibble(
      city   = city,
      time   = result$current$time,
      temp_f = result$current$temperature_2m
    )
  }) |>
  bind_rows()

current_df
```

::: {.callout-note}
## Reading the `map` Pattern

- `\(city, latitude, longitude) { ... }` is an **anonymous function** — a recipe with no name. The `\( )` lists the ingredients; the `{ }` holds the steps.
- `pmap()` runs the recipe once **per row** of `cities`.
- Each run returns a one-row tibble; `bind_rows()` stacks them into one data frame.

Same ETV steps as before — just repeated for us, without copy-paste and without defining named functions.
:::


#### ———————
### Q7. {.unnumbered}

**(Multiple Choice):** In the code above, what does `\(city, latitude, longitude)` mean?

a) It creates three new columns in the tibble.
b) It is an anonymous function — a nameless recipe whose inputs are one row's city, latitude, and longitude.
c) It deletes the variables city, latitude, and longitude.
d) It is a comment that R ignores.
e) It is a file path.

#### ———————


#### ———————
### Q8. {.unnumbered}

**(Fill in the Blank):** After `pmap()` returns a list of one-row tibbles, we stack them into a single data frame using \_\_\_\_\_\_\_\_.

#### ———————


---


## Part 6: Geocoding — Where Do the Coordinates Come From?

We typed the latitude and longitude by hand. But Open-Meteo also has a **geocoding** endpoint that converts a city *name* into coordinates. Browser first, as always — paste this:

```
https://geocoding-api.open-meteo.com/v1/search?name=San+Luis+Obispo&count=1
```

You'll see JSON with a `results` array whose first entry contains `latitude`, `longitude`, `population`, and more. Now the same thing in R — with `map()` and an anonymous function again:

```{r}
city_names <- c("San Luis Obispo", "Boston", "Ames")

coords_df <- city_names |>
  map(\(nm) {

    # E: extract
    result <- request("https://geocoding-api.open-meteo.com/v1/search") |>
      req_url_query(name = nm, count = 1) |>
      req_perform() |>
      resp_body_json()

    # T: transform (first hit in the results array)
    hit <- result$results[[1]]
    tibble(
      city      = nm,
      latitude  = round(hit$latitude, 2),
      longitude = round(hit$longitude, 2)
    )
  }) |>
  bind_rows()

coords_df
```

Now we can find *any* city's coordinates — which Session 4 will need, because the historical weather endpoint only speaks latitude/longitude.

::: {.callout-tip}
## 🎯 Goal Check
Notice we just ran the **same ETV pattern a third time** — different endpoint, same skills. That's the whole trick: every API interaction is the same cycle.
:::


#### ———————
### Q9. {.unnumbered}

**(Fill in the Blank):** The geocoding endpoint converts a city name into \_\_\_\_\_\_\_\_ and \_\_\_\_\_\_\_\_.

#### ———————


---


## Part 7: Your Turn — Build Your Own Request

Make a current weather request for **your hometown** (or any city you choose).

### Activity: Fill in the Blanks

```{r}
# Step 1: Find your city's coordinates with the geocoding endpoint
my_coords <- request("https://geocoding-api.open-meteo.com/v1/search") |>
  req_url_query(name = "________", count = 1) |>
  req_perform() |>
  resp_body_json()

my_lat <- my_coords$results[[1]]$________
my_lon <- my_coords$results[[1]]$________

# Step 2: Request the current weather (E)
my_result <- request("https://api.open-meteo.com/v1/________") |>
  req_url_query(
    latitude = ________,
    longitude = ________,
    current = "temperature_2m",
    temperature_unit = "fahrenheit"
  ) |>
  req_perform() |>
  resp_body_json()

# Step 3: Format the response into a tibble (T)
my_df <- tibble(
  city   = "________",
  time   = my_result$current$________,
  temp_f = my_result$current$________
)

my_df
```


#### ———————
### Q10. {.unnumbered}

**(Ordering — from Session 2):** Put the following steps in the correct order for making an API request in R. Write the letters in order.

a) Parse the JSON response into a data frame
b) Construct the URL with endpoint and query parameters
c) Check the status code for success (200)
d) Send the request using `req_perform()`
e) Create a request object with `request()`

#### ———————


---


## Session 3 Recap

Today the goal became real:

1. **Open-Meteo API** — free, no API key, endpoint `https://api.open-meteo.com/v1/forecast`
2. **Browser first** — a request is just a URL; the browser shows the raw JSON response
3. **ETV Cycle** — Extract (request), Transform (JSON → tibble), Visualize (Session 4)
4. **Extract in R** — `request()` |> `req_url_query()` |> `req_perform()` |> `resp_body_json()`
5. **Transform in R** — pulling values from the named list into a `tibble()`
6. **Scaling with `map()`** — anonymous functions `\( )` apply the same recipe to many cities, no copy-paste
7. **Geocoding** — converting city names into the coordinates other endpoints need

**Up next in Session 4:** the **V** in ETV — historical weather data for Boston vs. San Luis Obispo, and your own visualizations!


---


## During the Break {.unnumbered}

Before Session 4 starts:

- **Questions?** Come up and ask — Meg and I will be here for the whole break.
- **Did you finish?** Check yourself:
    - [ ] I pasted the forecast URL into my browser and saw JSON
    - [ ] I ran the R request and got a `200` status
    - [ ] I built `my_df` for my own city in Part 7
- If you finished everything, try adding a fourth city to the `cities` table in Part 5 and re-running the `pmap()` block.
