---
title: "Session 4: Historical Weather, Visualization & Wrap-Up"
subtitle: "Part B — Friday, August 7, 10:55–11:50 AM · Columbus I & II"
author: "Immanuel Williams"
date: "August 7, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    code-tools: true
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## Recap from Session 3

In Session 3 (*Weather API — First Real Requests*), we:

- Made our first real requests to the free **Open-Meteo** API (no API key!)
- Saw the request in the **browser first**, then automated it in R
- Learned the **ETV Cycle**: Extract → Transform → Visualize
- Used `map()` + anonymous functions `\( )` to repeat the recipe for many cities
- Used the **geocoding** endpoint to turn city names into coordinates

Today we complete the cycle: **historical** weather data and the **V** — visualization. As in Session 3, the code is **pre-built** — you run it, read it, and fill in the blanks (city names, dates, column names).

::: {.callout-important}
## 🎯 The One Goal — Final Step

> **Make a request to an API, then format the response into something we can analyze.**

Today "analyze" becomes concrete: real plots comparing real cities.
:::

### Load Libraries

```{r}
#| output: false
library(httr2)       # E: makes web requests
library(tibble)      # T: tidyverse data frames
library(dplyr)       # T: data manipulation
library(purrr)       # T: map() — apply a recipe to many inputs
library(ggplot2)     # V: visualizations
```


---


## Part 1: Historical Weather — The Full ETV Pipeline

### Look First

Inspect the historical request before running it:

```
https://archive-api.open-meteo.com/v1/archive?latitude=35.28&longitude=-120.66&start_date=2025-07-01&end_date=2025-07-07&daily=temperature_2m_max,temperature_2m_min&temperature_unit=fahrenheit&timezone=auto
```

| Component | Value | Purpose |
|---|---|---|
| Endpoint | `https://archive-api.open-meteo.com/v1/archive` | Historical-weather service |
| `latitude` | `35.28` | North–south location |
| `longitude` | `-120.66` | East–west location |
| `start_date` | `2025-07-01` | First date |
| `end_date` | `2025-07-07` | Last date |
| `daily` | `temperature_2m_max,temperature_2m_min` | Requested variables |
| `temperature_unit` | `fahrenheit` | Temperature unit |
| `timezone` | `auto` | Local dates and times |

### Now in Words

The endpoint is `https://archive-api.open-meteo.com/v1/archive` — a third door, following the same convention as Session 3's forecast and geocoding doors. Two new query parameters do the date work: `start_date` and `end_date`. And here's the good news — **one request returns the entire date range**. No loops needed.

Now the **full pipeline in R, pre-built**: geocode the city name → request the historical data → build the data frame. In your workshop file, the only blanks are the **city name** and the **two dates** — everything structural is written for you:

```{r}
# Step 0 — Geocode: city name → coordinates (Session 3's endpoint)
city_name <- "________"

geo <- request("https://geocoding-api.open-meteo.com/v1/search") |>
  req_url_query(name = city_name, count = 1) |>
  req_perform() |>
  resp_body_json()

hit <- geo$results[[1]]
my_lat <- round(hit$latitude, 2)
my_lon <- round(hit$longitude, 2)

# E: extract — historical request ---------------------------
result <- request("https://archive-api.open-meteo.com/v1/archive") |>
  req_url_query(
    latitude = my_lat,
    longitude = my_lon,
    start_date = "________",
    end_date = "________",
    daily = "temperature_2m_max,temperature_2m_min",
    temperature_unit = "fahrenheit",
    timezone = "auto"
  ) |>
  req_perform() |>
  resp_body_json()

# Peek at the daily piece
str(result$daily, max.level = 1)
```

The `daily` element is a named list of three parallel lists: `time`, `temperature_2m_max`, and `temperature_2m_min` — one entry per day.

```{r}
# T: transform ------------------------------------------------
slo_df <- tibble(
  city     = city_name,
  date     = as.Date(unlist(result$daily$time)),
  temp_max = unlist(result$daily$temperature_2m_max),
  temp_min = unlist(result$daily$temperature_2m_min)
)

slo_df
```

::: {.callout-note}
## Why `unlist()`?
`resp_body_json()` parses each JSON array into an R **list**. `unlist()` flattens that list into a plain **vector**, which is what a tibble column needs. List in, vector out — that's all it does.
:::


#### ———————
### Q1. {.unnumbered}

**(Fill in the Blank):** To ask the archive endpoint for a different range of days, we change the \_\_\_\_\_\_\_\_ and \_\_\_\_\_\_\_\_ query parameters.

#### ———————


#### ———————
### Q2. {.unnumbered}

**(Multiple Choice):** What does `unlist()` do in the Transform step?

a) It flattens a list into a plain vector so it can be a tibble column.
b) It sends the request to the server.
c) It removes missing values from the data.
d) It converts Fahrenheit to Celsius.
e) It sorts the dates in order.

#### ———————


---


## Part 2: Instructor Visualization Examples

### Example 1 — Highs and lows for one city

**Let’s visualize what we have already done.**

```{r}
# V: visualize -------------------------------------------------
ggplot(slo_df, aes(x = date)) +
  geom_line(aes(y = ________, color = "High"), linewidth = 1.2) +
  geom_line(aes(y = ________, color = "Low"), linewidth = 1.2) +
  scale_color_manual(values = c("High" = "red", "Low" = "blue")) +
  labs(
    title = "Daily High and Low Temperatures in San Luis Obispo",
    subtitle = "July 1–7, 2025 (Open-Meteo archive)",
    x = "Date",
    y = "Temperature (°F)",
    color = "Temperature"
  ) +
  theme_minimal(base_size = 14)
```

### Now in Words

Each `geom_line()` pulls one column from the tibble: `temp_max` becomes the red "High" line, `temp_min` the blue "Low" line, and `date` runs along the x-axis. Change a column name and the plot changes with it — the plot is just the data frame, drawn.

🎉 **That's the entire workshop goal, end to end:** we made a request, formatted the response, and analyzed it with a plot. Everything from here is variations on this cycle.


#### ———————
### Q3. {.unnumbered}

**(Fill in the Blank):** In the ETV cycle, `ggplot()` is the tool for the \_\_\_\_\_\_\_\_ step.

#### ———————


---


### Example 2 — Compare three cities

Now compare one measurement across three cities.

```{r}
cities <- tribble(
  ~city,             ~latitude, ~longitude,
  "San Luis Obispo",     35.28,    -120.66,
  "Boston",              42.36,     -71.06,
  "Ames",                42.03,     -93.62
)
```

#### Comparison 1: Summer

```{r}
# E + T for both cities: July 2025
summer_df <- cities |>
  pmap(\(city, latitude, longitude) {

    result <- request("https://archive-api.open-meteo.com/v1/archive") |>
      req_url_query(
        latitude = latitude,
        longitude = longitude,
        start_date = "2025-07-01",
        end_date = "2025-07-31",
        daily = "temperature_2m_max",
        temperature_unit = "fahrenheit",
        timezone = "auto"
      ) |>
      req_perform() |>
      resp_body_json()

    tibble(
      city     = city,
      date     = as.Date(unlist(result$daily$time)),
      temp_max = unlist(result$daily$temperature_2m_max)
    )
  }) |>
  bind_rows()

# V: one line per city
ggplot(summer_df, aes(x = date, y = temp_max, color = city)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Summer Daily Highs: Boston vs. San Luis Obispo",
    subtitle = "July 2025",
    x = "Date",
    y = "Daily High (°F)",
    color = "City"
  ) +
  theme_minimal(base_size = 14)
```


#### ———————
### Q4 — Visual Check (Summer). {.unnumbered}

**(Multiple Choice):** In the summer plot, what distinguishes Boston's line from San Luis Obispo's line?

a) The line type (solid vs. dashed)
b) The x-axis position
c) Separate panels for each city
d) The point shape
e) The line color, mapped with `aes(color = city)`

#### ———————


## Part 3: Independent Three-City Comparison

### Return to the motivating question

Build the same three-city historical comparison used in the Shiny capstone.

<details>
<summary>Show city, variable, and date choices</summary>

| 🏙 City (lat, lon) | 📊 Daily variable | 📅 Date range |
|---|---|---|
| Chicago (41.88, -87.63) | `temperature_2m_max` — daily high | Summer: `2025-06-01` → `2025-08-31` |
| Miami (25.76, -80.19) | `temperature_2m_min` — daily low | Winter: `2025-12-01` → `2026-02-28` |
| Denver (39.74, -104.98) | `precipitation_sum` — total precipitation | Spring: `2026-03-01` → `2026-05-31` |
| Seattle (47.61, -122.33) | `windspeed_10m_max` — peak wind | Last 3 months: `2026-04-01` → `2026-06-30` |
| Ames (42.03, -93.62) | `snowfall_sum` — snowfall | One month of your choice |
| *Your hometown (geocode it!)* | | |

</details>

### Activity: Fill in the Blanks

The pipeline is **pre-built** — your blanks are the city rows, the two dates, the daily variable, and the plot labels:

```{r}
# Choose exactly three cities, matching the Shiny comparison
my_cities <- tribble(
  ~city,        ~latitude, ~longitude,
  "________",    ________,   ________,
  "________",    ________,   ________,
  "________",    ________,   ________
)

# E + T: the same pre-built recipe
my_df <- my_cities |>
  pmap(\(city, latitude, longitude) {

    result <- request("https://archive-api.open-meteo.com/v1/archive") |>
      req_url_query(
        latitude = latitude,
        longitude = longitude,
        start_date = "________",
        end_date = "________",
        daily = "________",
        temperature_unit = "fahrenheit",
        timezone = "auto"
      ) |>
      req_perform() |>
      resp_body_json()

    tibble(
      city  = city,
      date  = as.Date(unlist(result$daily$time)),
      value = unlist(result$daily$________)
    )
  }) |>
  bind_rows()

# V: one line per city
ggplot(my_df, aes(x = date, y = value, color = city)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "________",
    x = "Date",
    y = "________",
    color = "City"
  ) +
  theme_minimal(base_size = 14)
```


---


## Part 4: Reflection, Classroom Applications & Survey

#### ———————
### Q5. {.unnumbered}

**(Discussion):** You just built a complete data pipeline: API request → JSON parsing → data wrangling → visualization. Think about the courses you teach. Where could API data naturally integrate into your curriculum? What topics could benefit from live, real-world data?

#### ———————


### Common Pitfalls When Working with APIs

1. **Rate Limits:** Most APIs limit how many requests you can make per minute/hour (Open-Meteo's free tier allows ~10,000 calls/day — generous, but not infinite). Design your code to respect limits.

2. **API Keys (elsewhere):** Most commercial APIs require keys. When you use one, never write the key directly in your script — store it in an environment file. The Practice Labs on this site walk through this.

3. **Data Structure Changes:** API providers can change their response format without warning. Always inspect the structure of new responses (`str()`) before building pipelines around them.

4. **URL Encoding:** Special characters (spaces, ampersands, accents) must be encoded in URLs. `req_url_query()` handles this for you automatically — one more reason to like it.

5. **Nested JSON Complexity:** Real-world JSON can be deeply nested. Practice with `str()` and `names()` to navigate the structure before trying to flatten it.


### Reflection Questions for Your Practice

- Where could API data naturally integrate into your curriculum?
- What are the pitfalls (rate limits, authentication) that students need to know?
- How does working with live, messy data differ from working with pre-cleaned datasets?
- What ethical considerations arise when collecting data from APIs?


### Workshop Summary

| Session | Day | Time | Key Topics |
|---------|-----|------|------------|
| 1 — API Mechanics & Response Fundamentals | Part A | 8:00–8:55 AM | API responses |
| 2 — The Anatomy of a Request | Part A | 8:55–9:50 AM | Requests and URLs |
| 3 — Weather API — First Real Requests | Part B | 10:00–10:55 AM | Current weather by city |
| 4 — Historical Weather, Visualization & Wrap-Up | Part B | 10:55–11:50 AM | Historical city comparisons |

**Key Takeaways:**

1. APIs provide **programmatic access** to live data
2. Requests are **URLs** — responses are **JSON**
3. Status codes help you **diagnose problems**
4. The **ETV cycle** (Extract → Transform → Visualize) is the same for every API
5. `map()` + anonymous functions repeat the recipe **without copy-paste**
6. Live data creates **authentic learning experiences**

Thank you for participating! Take these skills and bring real-world data into your classrooms.

::: {.callout-important}
## 📋 Before You Leave — MAA MathFest 2026 Workshop Survey

Please take a few minutes to complete the official workshop evaluation — it directly shapes future MAA MathFest workshops:

👉 **[MAA MathFest 2026 Evaluation / Survey](https://docs.google.com/forms/d/e/1FAIpQLSd2JNOT09AWYlI-5S8q4QyMTYsV0GKn0AV7R921rexumXgHuw/viewform?usp=dialog)**

Thank you — your feedback matters!
:::


---


## After the Workshop {.unnumbered}

- **Questions?** Come up and ask — Meg and I will stick around.
- **Did you finish?** Check yourself:
    - [ ] I ran the historical pipeline and plotted San Luis Obispo's July temperatures
    - [ ] I made both Boston vs. SLO comparison plots
    - [ ] I built and plotted **my own 2–3 city pipeline** in Part 3
    - [ ] I filled out the **workshop survey** (link above)
- **Resources for when this class is over:**
    - **Practice Labs on this site** — five guided labs (NASA, Cat API, OpenWeather, TMDB, GitHub), including APIs that *do* require keys
    - ****Take-Home Exercises** page on the workshop website** — dummy-server practice with GET/POST in R and error handling, plus function-building practice
    - [Open-Meteo documentation](https://open-meteo.com/en/docs) — every variable and endpoint we used, and many more
    - [httr2 package site](https://httr2.r-lib.org/) — the full request/response toolkit
    - [R for Data Science (2e), Iteration chapter](https://r4ds.hadley.nz/iteration.html) — more on `map()` and friends
    - [JSONPlaceholder](https://jsonplaceholder.typicode.com/) — the free dummy server, for classroom practice
