---
title: "Session 4: Historical Weather, Visualization & Wrap-Up"
subtitle: "Day 2 — 10:00 AM to 11:00 AM"
author: "Immanuel Williams"
date: "August 6, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    code-tools: true
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## Recap from Session 3

In Session 3, we:

- Set up our API key securely using `.Renviron`
- Made our first real request to the OpenWeather Current Weather API
- Built a `geocode()` function to convert city names to coordinates
- Practiced making our own requests

Now we'll combine geocoding with the **One Call API** to get **historical weather data**, loop over multiple days, and create visualizations.


---


## Part 1: Getting Historical Weather Data

The OpenWeather **One Call API 3.0** provides historical daily summaries for a given latitude, longitude, and date. This requires coordinates — which is why we built the `geocode()` function!

### The Plan

1. Use the Geocoding API to get lat/lon for a city
2. Define a range of past dates
3. Loop over each date, calling the One Call API
4. Combine all results into a single data frame

### Load Libraries

```{r}
#| output: false
library(httr2)
library(tibble)
library(lubridate)
library(ggplot2)
library(dplyr)
library(dotenv)
library(glue)
```

```{r}
#| results: hide
dotenv::load_dot_env(file = ".Renviron.txt")
```


### Step-by-Step: Past 5 Days of Weather

```{r}
# Step 1: Choose city and number of days
city_name <- "San Luis Obispo"
numdays <- 5
```

```{r}
# Step 2: Geocode the city
geo_url <- glue(
  "http://api.openweathermap.org/geo/1.0/direct?",
  "q=", URLencode(city_name),
  "&limit=1&appid=", Sys.getenv("API_KEY")
)

geo_response <- req_perform(request(geo_url))

if (resp_status(geo_response) == 200) {
  geo_data_df <- resp_body_json(geo_response) |>
    as.data.frame()
  
  if (length(geo_data_df) == 0) {
    stop("City not found. Please check the city name.")
  }
  
  lat <- round(geo_data_df$lat, digits = 2)
  lon <- round(geo_data_df$lon, digits = 2)
  
  cat("Coordinates for", city_name, "-> Lat:", lat, "Lon:", lon, "\n")
}
```

```{r}
# Step 3: Define date range
date_range <- as.character(lubridate::today() - days(1:numdays))
print(date_range)
```

::: {.callout-note}
## Why `as.character()`?
Without wrapping in `as.character()`, dates will mathematically evaluate and become integers when used in URL construction. We need them as strings like `"2026-08-05"` for the API.
:::

```{r}
# Step 4: Loop over dates and call the API
hist_weather_df <- data.frame()

for (date in date_range) {
  
  weather_url <- glue(
    "https://api.openweathermap.org/data/3.0/onecall/day_summary?",
    "lat=", lat,
    "&lon=", lon,
    "&date=", date,
    "&appid=", Sys.getenv("API_KEY"),
    "&units=imperial"
  )
  
  weather_response <- req_perform(request(weather_url))
  
  if (resp_status(weather_response) == 200) {
    daily_weather_df <- resp_body_json(weather_response) |>
      as.data.frame()
    
    # Add city and date columns
    daily_weather_df <- daily_weather_df |>
      mutate(
        city = city_name,
        date = date
      )
    
    hist_weather_df <- bind_rows(hist_weather_df, daily_weather_df)
    
  } else {
    warning(paste("Failed to get weather for", date, "-", resp_status(weather_response)))
  }
}

print(hist_weather_df)
```


### Understanding the Response Structure

The API returns a JSON object representing summary statistics for a specific location and date. The nested structure flattens into descriptive column names:

| JSON Key | R Column Name | Type | Description |
|----------|--------------|------|-------------|
| `temperature.min` | `temperature.min` | numeric | Minimum temp for the day |
| `temperature.max` | `temperature.max` | numeric | Maximum temp for the day |
| `wind.max.speed` | `wind.max.speed` | numeric | Peak wind speed |
| `date` (added) | `date` | character | From our loop variable |
| `city` (added) | `city` | character | City name from geocoding |

Example row:

| temperature.min | temperature.max | wind.max.speed | date | city |
|----------------|----------------|---------------|------|------|
| 50.1 | 75.6 | 12.3 | 2026-08-04 | San Luis Obispo |


#### ———————
### Q1. {.unnumbered}

**(Open-Ended):** In the loop above, we use `bind_rows()` to combine data from each day. Why do we initialize `hist_weather_df <- data.frame()` *before* the loop? What would happen if we didn't?

::: {.callout-tip collapse="true" title="Answer Q1"}
We initialize an empty data frame before the loop so that `bind_rows()` has something to append to on the first iteration. Without it, R would throw an error because `hist_weather_df` wouldn't exist when we try to bind the first day's results to it. This is a common pattern in R: initialize → loop → accumulate.
:::
#### ———————


#### ———————
### Q2. {.unnumbered}

**(Multiple Choice — Callback to Session 1 Q5):** In the loop, we use `warning()` instead of `stop()` when a single day's request fails. Why?

a) `warning()` is faster than `stop()`
b) `warning()` lets the loop continue to the next date, while `stop()` would halt everything
c) `warning()` sends a message to the API server
d) There is no practical difference

::: {.callout-tip collapse="true" title="Answer Q2"}
**b)** Using `warning()` lets the loop continue processing remaining dates even if one fails. If we used `stop()`, the entire function would halt at the first failure — losing all data for subsequent dates.
:::
#### ———————


---


## Part 2: Building a Reusable Function

Let's wrap the logic above into a clean, reusable function:

```{r}
prev_weather <- function(city, days) {
  
  # --- Geocoding ---
  geo_url <- glue(
    "http://api.openweathermap.org/geo/1.0/direct?",
    "q=", URLencode(city),
    "&limit=1&appid=", Sys.getenv("API_KEY")
  )
  
  geo_response <- req_perform(request(geo_url))
  
  if (resp_status(geo_response) != 200) {
    stop("Geocoding failed. Check your API key or city name.")
  }
  
  geo_data_df <- resp_body_json(geo_response) |> as.data.frame()
  
  if (length(geo_data_df) == 0) {
    stop("City not found. Please check the city name.")
  }
  
  lat <- round(geo_data_df$lat, digits = 2)
  lon <- round(geo_data_df$lon, digits = 2)
  cat("Coordinates for", city, "-> Lat:", lat, "Lon:", lon, "\n")
  
  # --- Historical Weather ---
  date_range <- as.character(lubridate::today() - days(1:days))
  hist_weather_df <- data.frame()
  
  for (date in date_range) {
    weather_url <- glue(
      "https://api.openweathermap.org/data/3.0/onecall/day_summary?",
      "lat=", lat,
      "&lon=", lon,
      "&date=", date,
      "&appid=", Sys.getenv("API_KEY"),
      "&units=imperial"
    )
    
    weather_response <- req_perform(request(weather_url))
    
    if (resp_status(weather_response) == 200) {
      daily_weather_df <- resp_body_json(weather_response) |>
        as.data.frame() |>
        mutate(city = city, date = date)
      
      hist_weather_df <- bind_rows(hist_weather_df, daily_weather_df)
    } else {
      warning(paste("Failed to get weather for", date, "-", resp_status(weather_response)))
    }
  }
  
  return(hist_weather_df)
}
```


### Test the Function

```{r}
# Get past 4 days of weather for your city
prev_weather("San Luis Obispo", 4)
```

```{r}
# Try another city!
prev_weather("Ames, IA, USA", 4)
```


#### ———————
### Q3. {.unnumbered}

**(Open-Ended — Callback to Session 2 Q8):** In Session 2, we outlined the steps for making an API request: construct URL → create request → perform → check status → parse. How does the `prev_weather` function follow this same pattern, but extend it with a loop?

::: {.callout-tip collapse="true" title="Answer Q3"}
The `prev_weather` function follows the exact same 5-step pattern we learned, but runs it **twice**: once for geocoding (single request) and then **multiple times** inside a loop for each day's weather data. Each iteration constructs a new URL (with a different date), creates a request, performs it, checks the status, and parses the result. The added complexity is the accumulation step — binding each day's data frame into a growing combined result.
:::
#### ———————


---


## Part 3: Visualizing Weather Data

Now let's make the data come alive with visualizations!

### Temperature Over Time

```{r}
#| message: false

# Get data
df4 <- prev_weather("Ames, IA, USA", 4) |>
  mutate(date = as.Date(date))

# Plot high and low temperatures
ggplot(df4, aes(x = date)) +
  geom_line(aes(y = temperature.max, color = "High"), linewidth = 1.2) +
  geom_line(aes(y = temperature.min, color = "Low"), linewidth = 1.2) +
  scale_color_manual(values = c("High" = "red", "Low" = "blue")) +
  labs(
    title = paste("High and Low Temperatures in", unique(df4$city)),
    x = "Date",
    y = "Temperature (°F)",
    color = "Temperature Type"
  ) +
  theme_minimal(base_size = 14)
```


### Multi-City Comparison Map

Let's compare weather across multiple cities using a map:

```{r}
#| output: false
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

```{r}
cities <- c("San Luis Obispo", "Santa Barbara", "San Francisco")

# Function to get coordinates + weather summary for a single city
get_city_weather <- function(city, date = Sys.Date()) {
  geo_url <- glue(
    "http://api.openweathermap.org/geo/1.0/direct?",
    "q=", URLencode(city),
    "&limit=1&appid=", Sys.getenv("API_KEY")
  )
  
  geo_response <- req_perform(request(geo_url))
  
  if (resp_status(geo_response) == 200) {
    geo_data_df <- as.data.frame(resp_body_json(geo_response))
    
    if (length(geo_data_df) == 0) return(NULL)
    
    lat <- geo_data_df$lat
    lon <- geo_data_df$lon
    
    weather_url <- glue(
      "https://api.openweathermap.org/data/3.0/onecall/day_summary?",
      "lat=", lat,
      "&lon=", lon,
      "&date=", format(date, "%Y-%m-%d"),
      "&appid=", Sys.getenv("API_KEY"),
      "&units=imperial"
    )
    
    weather_response <- req_perform(request(weather_url))
    
    if (resp_status(weather_response) == 200) {
      weather_data <- resp_body_json(weather_response)
      
      tibble(
        city = city,
        date = date,
        lat = lat,
        lon = lon,
        temp_max = weather_data$temperature$max,
        temp_min = weather_data$temperature$min
      )
    } else return(NULL)
  } else return(NULL)
}

# Fetch for all cities
weather_df <- bind_rows(lapply(cities, get_city_weather))
```

```{r}
# Convert to spatial data
weather_sf <- weather_df |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

# Get California boundary
ca <- ne_states(country = "United States of America", returnclass = "sf") |>
  filter(name == "California")

# Plot
ggplot() +
  geom_sf(data = ca, fill = "gray95", color = "gray60") +
  geom_sf(data = weather_sf, aes(color = temp_max), size = 6) +
  geom_sf_label(data = weather_sf, aes(label = city), nudge_y = 0.5, size = 2.5) +
  scale_color_viridis_c(option = "plasma", name = "High Temp (°F)") +
  labs(
    title = "California City Temperatures",
    subtitle = paste("Date:", Sys.Date())
  ) +
  theme_minimal()
```


#### ———————
### Q4. {.unnumbered}

**(Open-Ended):** We used `lapply(cities, get_city_weather)` to apply the function to multiple cities. How is this different from using a `for` loop? When might you prefer one over the other?

::: {.callout-tip collapse="true" title="Answer Q4"}
`lapply()` applies a function to each element of a list and returns a list of results — it's a more "functional programming" approach. A `for` loop does the same thing but requires manually initializing and accumulating results. `lapply()` is more concise and idiomatic in R. However, `for` loops can be easier to read and debug, especially when you need to handle errors or track progress. For API work with potential failures, a `for` loop with `tryCatch` or `warning()` can be more transparent.
:::
#### ———————


---


## Part 4: Your Turn — Final Activity

### Activity: Build Your Own Weather Report

Using everything we've learned across all four sessions, complete the following tasks:

**Task 1:** Choose **3 cities** that are meaningful to you (hometown, current city, dream destination).

```{r}
# Your cities here
my_cities <- c("_________", "_________", "_________")
```

**Task 2:** Use the `prev_weather()` function to get the past 3 days of weather for one of your cities.

```{r}
# Get weather data
my_weather <- prev_weather("_________", 3)
print(my_weather)
```

**Task 3:** Create a visualization of the temperature data.

```{r}
# Your visualization here
my_weather <- my_weather |> mutate(date = as.Date(date))

ggplot(my_weather, aes(x = date)) +
  geom_line(aes(y = temperature.max, color = "High"), linewidth = 1.2) +
  geom_line(aes(y = temperature.min, color = "Low"), linewidth = 1.2) +
  scale_color_manual(values = c("High" = "red", "Low" = "blue")) +
  labs(
    title = paste("Temperature in", unique(my_weather$city)),
    x = "Date",
    y = "Temperature (°F)",
    color = ""
  ) +
  theme_minimal(base_size = 14)
```

**Bonus Task:** Modify the `get_city_weather()` function to also retrieve `humidity` or `wind.max.speed` and add it to the tibble output.


#### ———————
### Q5. {.unnumbered}

**(Discussion):** You just built a complete data pipeline: API request → JSON parsing → data wrangling → visualization. Think about the courses you teach. Where could API data naturally integrate into your curriculum? What topics could benefit from live, real-world data?

::: {.callout-tip collapse="true" title="Example Answers"}
Possible integration points:

- **Intro Stats:** Use weather data for descriptive statistics (mean temperature, variability across cities)
- **Regression:** Model temperature as a function of latitude or time
- **Data Wrangling:** Parse nested JSON, clean and reshape API responses
- **Probability:** Explore weather prediction accuracy (forecast vs. actual)
- **Capstone Projects:** Let students choose an API and build their own data pipeline
:::
#### ———————


#### ———————
### Q6. {.unnumbered}

**(Open-Ended — Final Callback):** Across all four sessions, we've covered JSON, request/response, CRUD, GET/POST, URL anatomy, status codes, API keys, geocoding, loops, and visualization. Which concept was most surprising or new to you, and how might it change how you approach data collection in your work?

::: {.callout-tip collapse="true" title="Answer Q6"}
This is a personal reflection question — there's no single right answer. The goal is to identify which concept shifted your thinking and to consider practical applications in your teaching or research.
:::
#### ———————


---


## Part 5: Reflection & Pitfalls

### Common Pitfalls When Working with APIs

1. **Rate Limits:** Most APIs limit how many requests you can make per minute/hour. The free OpenWeather tier allows 60 calls/minute. Design your code to respect these limits.

2. **Authentication Expiry:** API keys can expire or be revoked. Always check your key status if you get `401` errors.

3. **Data Structure Changes:** API providers can change their response format without warning. Always inspect the structure of new responses before building pipelines around them.

4. **URL Encoding:** Special characters (spaces, ampersands, accents) must be encoded. Use `URLencode()`.

5. **Nested JSON Complexity:** Real-world JSON can be deeply nested. Practice with `str()` and `names()` to navigate the structure before trying to flatten it.


### Reflection Questions for Your Practice

- Where could API data naturally integrate into your curriculum?
- What are the pitfalls (rate limits, authentication) that students need to know?
- How does working with live, messy data differ from working with pre-cleaned datasets?
- What ethical considerations arise when collecting data from APIs?


---


## Workshop Summary

| Session | Day | Time | Key Topics |
|---------|-----|------|------------|
| 1 | Day 1 | 8–9 AM | JSON fundamentals, request/response, status codes |
| 2 | Day 1 | 9–10 AM | CRUD, GET/POST, URL anatomy, dummy server practice |
| 3 | Day 2 | 9–10 AM | API key setup, current weather, geocoding functions |
| 4 | Day 2 | 10–11 AM | Historical weather, loops, visualization, reflection |

**Key Takeaways:**

1. APIs provide **programmatic access** to live data
2. Requests are **URLs** — responses are **JSON**
3. Status codes help you **diagnose problems**
4. Functions make your API code **reusable and clean**
5. Live data creates **authentic learning experiences**

Thank you for participating! Take these skills and bring real-world data into your classrooms.
