---
title: "Practice Lab 1: NASA Open APIs"
subtitle: "Astronomy Picture of the Day & Near-Earth Object Data"
author: "Immanuel Williams"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    css: ../../custom.css
    code-fold: false
    code-tools: true
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

NASA provides free, production-quality APIs used by researchers worldwide. In this lab you will work with two:

- **APOD** — Astronomy Picture of the Day: one endpoint, one JSON object, a clean introduction.
- **NEO** — Near-Earth Object Web Service: asteroid data with a deeply nested response structure that requires real parsing work.

**Auth pattern:** API key as a **URL query parameter** (`?api_key=YOUR_KEY`).

---

## Step 1 — Get Your API Key

1. Go to [api.nasa.gov](https://api.nasa.gov)
2. Fill in the short registration form and submit
3. Your key arrives by email within minutes
4. Add it to your `.Renviron` file:

```
NASA_KEY=your_key_here
```

::: {.callout-note}
## DEMO_KEY — No signup needed
NASA provides a built-in demo key. Replace `nasa_key` with `"DEMO_KEY"` anywhere in this lab. It allows 30 requests/hour and 50/day — enough to complete the exercises.
:::

---

## Step 2 — Setup

```{r load-libs}
library(httr2)
library(dplyr)
library(ggplot2)
library(glue)
library(dotenv)

load_dot_env()
nasa_key <- Sys.getenv("NASA_KEY")
```

---

## Exercise 1: Astronomy Picture of the Day

The APOD endpoint returns a single JSON object describing today's featured image or video, along with a scientific explanation written by a professional astronomer.

```{r apod-request}
apod_url <- glue("https://api.nasa.gov/planetary/apod?api_key={nasa_key}")

req      <- request(apod_url)
response <- req_perform(req)

# Always check the status first
resp_status(response)
```

```{r apod-parse}
apod <- resp_body_json(response)

# Inspect the top-level keys
names(apod)
```

```{r apod-display}
cat("Title:     ", apod$title,      "\n")
cat("Date:      ", apod$date,       "\n")
cat("Media type:", apod$media_type, "\n")
cat("URL:       ", apod$url,        "\n")
cat("\nExplanation:\n", apod$explanation, "\n")
```

### Challenge A — Request a Specific Date

Add a `date` query parameter to retrieve any past APOD. Dates go back to 1995-06-16.

```{r apod-date}
apod_dated_url <- glue(
  "https://api.nasa.gov/planetary/apod",
  "?api_key={nasa_key}&date=1995-06-16"
)

apod_first <- resp_body_json(req_perform(request(apod_dated_url)))
cat("First ever APOD:", apod_first$title, "\n")
```

### Challenge B — Retrieve a Batch of APODs

Use the `count` parameter to retrieve multiple random APODs in a single request.

```{r apod-batch}
apod_batch_url <- glue(
  "https://api.nasa.gov/planetary/apod",
  "?api_key={nasa_key}&count=5"
)

apod_list <- resp_body_json(req_perform(request(apod_batch_url)))

# How is the structure different from a single APOD?
length(apod_list)
apod_list[[1]]$title
```

---

## Exercise 2: Near-Earth Object (Asteroid) Data

The NEO endpoint returns every asteroid that passed near Earth within a date range. The response is nested three levels deep — a great exercise in list navigation.

```{r neo-request}
neo_url <- glue(
  "https://api.nasa.gov/neo/rest/v1/feed",
  "?start_date=2024-01-01&end_date=2024-01-07",
  "&api_key={nasa_key}"
)

response <- req_perform(request(neo_url))
neo_data <- resp_body_json(response)

cat("Total asteroids in this range:", neo_data$element_count, "\n")
cat("Top-level keys:", paste(names(neo_data), collapse=", "), "\n")
```

```{r neo-explore}
# Near-Earth objects are nested: neo_data$near_earth_objects → date → list of asteroids
dates <- names(neo_data$near_earth_objects)
cat("Dates in response:", paste(dates, collapse=", "), "\n")

# Look at the first asteroid on the first date
first_day   <- neo_data$near_earth_objects[[dates[1]]]
first_rock  <- first_day[[1]]

cat("Name:", first_rock$name, "\n")
cat("Hazardous?", first_rock$is_potentially_hazardous_asteroid, "\n")
cat("Min diameter (km):",
    first_rock$estimated_diameter$kilometers$estimated_diameter_min, "\n")
```

```{r neo-flatten}
# Flatten all dates and asteroids into a single data frame
all_asteroids <- lapply(neo_data$near_earth_objects, function(day) {
  lapply(day, function(rock) {
    data.frame(
      name          = rock$name,
      date          = rock$close_approach_data[[1]]$close_approach_date,
      is_hazardous  = rock$is_potentially_hazardous_asteroid,
      min_diam_km   = rock$estimated_diameter$kilometers$estimated_diameter_min,
      max_diam_km   = rock$estimated_diameter$kilometers$estimated_diameter_max,
      miss_dist_km  = as.numeric(rock$close_approach_data[[1]]$miss_distance$kilometers),
      speed_kmh     = as.numeric(rock$close_approach_data[[1]]$relative_velocity$kilometers_per_hour)
    )
  }) |> bind_rows()
}) |> bind_rows()

glimpse(all_asteroids)
```

```{r neo-plot}
ggplot(all_asteroids, aes(x = miss_dist_km / 1e6, fill = is_hazardous)) +
  geom_histogram(bins = 25, alpha = 0.85, color = "white") +
  scale_fill_manual(
    values = c("FALSE" = "#4f46e5", "TRUE" = "#ef4444"),
    labels = c("Not Hazardous", "Potentially Hazardous")
  ) +
  labs(
    title    = "Asteroid Miss Distances — Jan 1–7, 2024",
    subtitle = "Each bar = one asteroid; distance from Earth's center",
    x        = "Miss Distance (millions of km)",
    y        = "Count",
    fill     = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

---

## Questions

#### ———————
### Q1. {.unnumbered}

**(Conceptual):** In this lab the API key is passed as a query parameter (`?api_key=...`). Name one situation where this approach could create a security risk.

::: {.callout-tip collapse="true" title="Answer Q1"}
The key is visible in the URL — which means it appears in browser history, server access logs, and anywhere the URL is copy-pasted or shared. If you accidentally post the URL in a public place (a GitHub issue, a Slack message, a screenshot), your key is exposed. Header-based auth keeps the credential out of the URL entirely.
:::
#### ———————


#### ———————
### Q2. {.unnumbered}

**(Code):** The `count` parameter returns a list of APODs instead of a single object. Write code that extracts just the `title` from each item in the batch and prints them as a character vector.

::: {.callout-tip collapse="true" title="Answer Q2"}
```r
titles <- sapply(apod_list, function(x) x$title)
print(titles)
```
`sapply` maps a function over each list element and simplifies the result to a vector.
:::
#### ———————


#### ———————
### Q3. {.unnumbered}

**(Analysis):** After building `all_asteroids`, use `dplyr` to find the single asteroid with the **smallest miss distance** in the date range. Was it potentially hazardous?

::: {.callout-tip collapse="true" title="Answer Q3"}
```r
all_asteroids |>
  arrange(miss_dist_km) |>
  select(name, date, miss_dist_km, is_hazardous) |>
  slice(1)
```
The answer varies by date range, but notice whether `is_hazardous` is `TRUE` for the closest pass — this is the heart of what NASA tracks with NEO Watch.
:::
#### ———————
