---
title: "Practice Lab 3: OpenWeather Extended"
subtitle: "Air Quality, Forecasts & Multi-City Comparisons"
author: "Immanuel Williams"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    css: ../../custom.css
    code-fold: false
    code-tools: true
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

You already have your OpenWeather API key from **Session 3**. This lab takes you further into the same API — exploring endpoints you haven't used yet and building toward multi-city comparisons.

You will use:

- **Air Quality Index (AQI)** — requires coordinates (lat/lon), not just a city name
- **5-Day Forecast** — 40 time-step nested response across 5 days
- **Multi-city pipeline** — map a function across a vector of cities

**Auth pattern:** Same as Sessions 3–4 — API key as `appid` query parameter.

---

## Step 1 — Setup

```{r load-libs}
library(httr2)
library(dplyr)
library(ggplot2)
library(lubridate)
library(glue)
library(dotenv)

load_dot_env()
weather_key <- Sys.getenv("OPENWEATHER_KEY")
```

---

## Exercise 1: Air Quality Index

The AQI endpoint requires **latitude and longitude**, not a city name. This means you need to geocode first — exactly the pattern from Session 3.

### Step 1a — Geocode a city

```{r geocode}
get_coordinates <- function(city, key) {
  url <- glue("http://api.openweathermap.org/geo/1.0/direct?q={city}&limit=1&appid={key}")
  result <- resp_body_json(req_perform(request(url)))
  list(lat = result[[1]]$lat, lon = result[[1]]$lon)
}

coords <- get_coordinates("San Luis Obispo", weather_key)
cat("Lat:", coords$lat, "| Lon:", coords$lon, "\n")
```

### Step 1b — Fetch air quality

```{r aqi-request}
aqi_url <- glue(
  "http://api.openweathermap.org/data/2.5/air_pollution",
  "?lat={coords$lat}&lon={coords$lon}&appid={weather_key}"
)

response <- req_perform(request(aqi_url))
aqi_data <- resp_body_json(response)

# AQI scale: 1=Good, 2=Fair, 3=Moderate, 4=Poor, 5=Very Poor
aqi_value  <- aqi_data$list[[1]]$main$aqi
components <- aqi_data$list[[1]]$components

cat("Air Quality Index:", aqi_value, "\n")
cat("PM2.5:", components$pm2_5, "μg/m³\n")
cat("CO:",    components$co,    "μg/m³\n")
cat("NO2:",   components$no2,   "μg/m³\n")
cat("O3:",    components$o3,    "μg/m³\n")
```

---

## Exercise 2: 5-Day Forecast

The forecast endpoint returns 40 time steps — one every 3 hours over 5 days. The data lives inside a `list` key (an array).

```{r forecast-request}
forecast_url <- glue(
  "https://api.openweathermap.org/data/2.5/forecast",
  "?q=San+Luis+Obispo&appid={weather_key}&units=imperial"
)

response      <- req_perform(request(forecast_url))
forecast_data <- resp_body_json(response)

cat("Number of time steps:", length(forecast_data$list), "\n")
cat("City:", forecast_data$city$name, ",", forecast_data$city$country, "\n")
```

```{r forecast-flatten}
# Flatten the nested list into a tidy data frame
forecast_df <- lapply(forecast_data$list, function(step) {
  data.frame(
    datetime    = as_datetime(step$dt),
    temp        = step$main$temp,
    feels_like  = step$main$feels_like,
    humidity    = step$main$humidity,
    description = step$weather[[1]]$description,
    wind_speed  = step$wind$speed,
    pop         = step$pop  # probability of precipitation
  )
}) |> bind_rows()

glimpse(forecast_df)
```

```{r forecast-plot}
ggplot(forecast_df, aes(x = datetime, y = temp)) +
  geom_line(color = "#4f46e5", linewidth = 1) +
  geom_ribbon(aes(ymin = temp - 3, ymax = temp + 3), alpha = 0.15, fill = "#6366f1") +
  geom_point(aes(color = pop), size = 2) +
  scale_color_gradient(low = "#a5b4fc", high = "#ef4444",
                       name = "Rain Prob.") +
  labs(
    title    = "5-Day Temperature Forecast — San Luis Obispo",
    subtitle = "Point color indicates probability of precipitation",
    x        = NULL,
    y        = "Temperature (°F)"
  ) +
  theme_minimal(base_size = 13)
```

---

## Exercise 3: Compare Multiple Cities

Write a reusable function, then map it across a list of cities. This is the core pattern for any multi-source data pipeline.

```{r city-function}
get_current_weather <- function(city, key) {
  url <- glue(
    "https://api.openweathermap.org/data/2.5/weather",
    "?q={city}&appid={key}&units=imperial"
  )

  response <- req_perform(request(url))

  if (resp_status(response) != 200) {
    message("Failed for city: ", city)
    return(NULL)
  }

  d <- resp_body_json(response)

  data.frame(
    city        = d$name,
    country     = d$sys$country,
    temp        = d$main$temp,
    feels_like  = d$main$feels_like,
    humidity    = d$main$humidity,
    description = d$weather[[1]]$description,
    wind_speed  = d$wind$speed
  )
}
```

```{r city-map}
cities <- c(
  "San Luis Obispo,US",
  "Chicago,US",
  "Miami,US",
  "Seattle,US",
  "Denver,US"
)

# Map the function across all cities
weather_comparison <- lapply(cities, get_current_weather, key = weather_key) |>
  bind_rows()

print(weather_comparison)
```

```{r city-plot}
ggplot(weather_comparison, aes(x = reorder(city, temp), y = temp, fill = temp)) +
  geom_col(width = 0.6, radius = unit(4, "pt")) +
  geom_text(aes(label = paste0(round(temp), "°F")),
            hjust = -0.15, fontface = "bold", size = 4) +
  scale_fill_gradient(low = "#4f46e5", high = "#ef4444") +
  coord_flip() +
  labs(
    title = "Current Temperature by City",
    x     = NULL,
    y     = "Temperature (°F)",
    fill  = "Temp (°F)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

## Questions

#### ———————
### Q1. {.unnumbered}

**(Conceptual):** The AQI endpoint requires lat/lon, not a city name. Why do you think some API endpoints work with city names while others require coordinates?

::: {.callout-tip collapse="true" title="Answer Q1"}
City names are ambiguous — there are many cities named "Springfield" or "Paris." Coordinates are globally unique. Endpoints that measure physical phenomena at a precise geographic point (like air quality or elevation) require coordinates because a city-level approximation would be inaccurate. The geocoding step converts the human-readable name into the precise coordinates the scientific endpoint needs.
:::
#### ———————


#### ———————
### Q2. {.unnumbered}

**(Code):** The 5-day forecast `datetime` column comes from a Unix timestamp (`step$dt`). What does `as_datetime()` from `lubridate` do to it, and what would the raw value look like without that conversion?

::: {.callout-tip collapse="true" title="Answer Q2"}
`as_datetime()` converts the raw **Unix timestamp** (an integer counting seconds since January 1, 1970) into a human-readable `POSIXct` datetime object. Without conversion, the column would contain numbers like `1704067200` — correct but unreadable and unusable for time-based plotting.
:::
#### ———————


#### ———————
### Q3. {.unnumbered}

**(Extension):** Modify `get_current_weather()` to also retrieve the **air quality index** for each city. You'll need to geocode the city first, then call the AQI endpoint with the coordinates. What challenges do you run into?

::: {.callout-tip collapse="true" title="Answer Q3"}
The main challenge is **making two API calls per city** instead of one — geocoding (to get lat/lon) and then AQI (using those coordinates). This doubles the number of requests and makes rate limiting more of a concern. A robust solution caches the coordinates and handles errors at each step:

```r
get_weather_and_aqi <- function(city, key) {
  coords <- get_coordinates(city, key)
  weather <- get_current_weather(city, key)

  aqi_url  <- glue("http://api.openweathermap.org/data/2.5/air_pollution?lat={coords$lat}&lon={coords$lon}&appid={key}")
  aqi_resp <- resp_body_json(req_perform(request(aqi_url)))

  weather$aqi <- aqi_resp$list[[1]]$main$aqi
  weather
}
```
:::
#### ———————
