---
title: "Practice Lab 4: TMDB Movie Database"
subtitle: "API Keys, Bearer Tokens & Movie Data"
author: "Immanuel Williams"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    css: ../../custom.css
    code-fold: false
    code-tools: true
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

TMDB (The Movie Database) is a community-built database of movies, TV shows, actors, and genres. It is an excellent API for learning because it supports **two authentication methods simultaneously** — you'll use both.

- **API Key** — passed as a query parameter, same pattern as NASA
- **Bearer Token** — passed in the `Authorization` header, the industry-standard modern approach

Understanding Bearer tokens is critical: every major professional API (Spotify, Stripe, GitHub, Google Cloud) uses this pattern.

**Auth patterns:** API key in URL *and* Bearer token in header.

---

## Step 1 — Get Your Credentials

1. Go to [themoviedb.org](https://www.themoviedb.org) and create a free account
2. Go to **Settings → API** (left sidebar)
3. Request an API key — select "Developer" and fill in the form
4. On the same page you will see:
   - **API Key (v3)** — a 32-character hex string
   - **Read Access Token (v4 auth)** — a long JWT-style Bearer token
5. Add both to your `.Renviron`:

```
TMDB_KEY=your_api_key_here
TMDB_TOKEN=your_read_access_token_here
```

---

## Step 2 — Setup

```{r load-libs}
library(httr2)
library(dplyr)
library(ggplot2)
library(glue)
library(dotenv)

load_dot_env()
tmdb_key   <- Sys.getenv("TMDB_KEY")
tmdb_token <- Sys.getenv("TMDB_TOKEN")
```

---

## Exercise 1: Search Movies Using an API Key

This is the simplest auth pattern — identical to how you used the NASA and OpenWeather APIs.

```{r search-key}
req <- request("https://api.themoviedb.org/3/search/movie") |>
  req_url_query(
    api_key  = tmdb_key,
    query    = "Interstellar",
    language = "en-US",
    page     = 1
  )

response <- req_perform(req)
search_data <- resp_body_json(response)

cat("Total results:", search_data$total_results, "\n")
cat("Pages:", search_data$total_pages, "\n")
```

```{r search-flatten}
# Flatten the results array
movies_df <- lapply(search_data$results, function(m) {
  data.frame(
    id            = m$id,
    title         = m$title,
    release_date  = m$release_date %||% NA_character_,
    vote_average  = m$vote_average,
    vote_count    = m$vote_count,
    popularity    = m$popularity,
    overview      = substr(m$overview %||% "", 1, 100)
  )
}) |> bind_rows()

movies_df |> select(title, release_date, vote_average, popularity)
```

---

## Exercise 2: Get Popular Movies Using a Bearer Token

Now use the Bearer token. The endpoint is the same family, but the credential moves from the URL to an `Authorization` header.

```{r popular-bearer}
req <- request("https://api.themoviedb.org/3/movie/popular") |>
  req_headers(
    Authorization = glue("Bearer {tmdb_token}"),
    Accept        = "application/json"
  ) |>
  req_url_query(language = "en-US", page = 1)

response   <- req_perform(req)
pop_data   <- resp_body_json(response)

popular_df <- lapply(pop_data$results, function(m) {
  data.frame(
    title        = m$title,
    release_date = m$release_date %||% NA_character_,
    vote_average = m$vote_average,
    popularity   = m$popularity,
    genre_ids    = paste(unlist(m$genre_ids), collapse = ",")
  )
}) |> bind_rows()

head(popular_df, 10)
```

::: {.callout-important}
## Comparing the Two Approaches
Notice what changed: with the API key, credentials were in the URL (`?api_key=...`). With the Bearer token, the URL is clean and the credential is in the `Authorization` header. The response structure is identical — only the auth mechanism differs. This is exactly the difference you saw between the NASA lab and the Cat API lab.
:::

---

## Exercise 3: Retrieve Genres and Join

TMDB's results include `genre_ids` (numbers) rather than genre names. This exercise fetches the genre lookup table and joins it.

```{r genres}
req <- request("https://api.themoviedb.org/3/genre/movie/list") |>
  req_headers(Authorization = glue("Bearer {tmdb_token}")) |>
  req_url_query(language = "en-US")

genre_data <- resp_body_json(req_perform(req))

genre_df <- lapply(genre_data$genres, function(g) {
  data.frame(genre_id = g$id, genre_name = g$name)
}) |> bind_rows()

print(genre_df)
```

```{r genre-join}
# For each movie, get its primary genre (first in the list)
popular_df_genres <- popular_df |>
  mutate(primary_genre_id = as.integer(sub(",.*", "", genre_ids))) |>
  left_join(genre_df, by = c("primary_genre_id" = "genre_id"))

popular_df_genres |>
  count(genre_name, sort = TRUE)
```

```{r genre-plot}
popular_df_genres |>
  count(genre_name, sort = TRUE) |>
  filter(!is.na(genre_name)) |>
  ggplot(aes(x = reorder(genre_name, n), y = n, fill = n)) +
  geom_col(width = 0.7) +
  scale_fill_gradient(low = "#a5b4fc", high = "#4f46e5") +
  coord_flip() +
  labs(
    title = "Most Common Genres Among Currently Popular Movies",
    x     = NULL,
    y     = "Number of Movies",
    fill  = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

## Exercise 4: Paginate Through Results

TMDB returns 20 results per page. This exercise collects multiple pages.

```{r paginate}
get_popular_page <- function(page, token) {
  req <- request("https://api.themoviedb.org/3/movie/popular") |>
    req_headers(Authorization = glue("Bearer {token}")) |>
    req_url_query(language = "en-US", page = page)

  data <- resp_body_json(req_perform(req))

  lapply(data$results, function(m) {
    data.frame(
      title        = m$title,
      vote_average = m$vote_average,
      popularity   = m$popularity
    )
  }) |> bind_rows()
}

# Collect 3 pages (60 movies)
pages_1_to_3 <- lapply(1:3, get_popular_page, token = tmdb_token) |>
  bind_rows()

cat("Total movies collected:", nrow(pages_1_to_3), "\n")
```

---

## Questions

#### ———————
### Q1. {.unnumbered}

**(Conceptual):** You used two credentials in this lab: an API Key and a Bearer token. From TMDB's settings page, both credentials grant the same read access. What advantage does the Bearer token approach provide over the API key approach?

::: {.callout-tip collapse="true" title="Answer Q1"}
The Bearer token keeps the credential **out of the URL** — it travels in the Authorization header instead. This means it won't appear in server access logs, browser history, or accidental copy-pastes of URLs. Bearer tokens are also the foundation of OAuth 2.0, which enables more sophisticated flows like expiring tokens and scoped permissions — making them the preferred standard for professional APIs.
:::
#### ———————


#### ———————
### Q2. {.unnumbered}

**(Code):** TMDB paginates results with a `page` parameter. Write a function that retrieves **all pages** of a search query (not just page 1), using the `total_pages` value from the first response to know when to stop.

::: {.callout-tip collapse="true" title="Answer Q2"}
```r
search_all_pages <- function(query, key) {
  first_resp <- resp_body_json(req_perform(
    request("https://api.themoviedb.org/3/search/movie") |>
      req_url_query(api_key = key, query = query, page = 1)
  ))

  total_pages <- min(first_resp$total_pages, 5)  # cap at 5 to be polite

  all_pages <- lapply(seq_len(total_pages), function(p) {
    data <- resp_body_json(req_perform(
      request("https://api.themoviedb.org/3/search/movie") |>
        req_url_query(api_key = key, query = query, page = p)
    ))
    lapply(data$results, function(m) data.frame(title = m$title, vote_average = m$vote_average)) |>
      bind_rows()
  }) |> bind_rows()

  all_pages
}
```
Capping at 5 pages avoids hammering the API — always be a polite client.
:::
#### ———————


#### ———————
### Q3. {.unnumbered}

**(Analysis):** Using `popular_df_genres`, find the top 5 most popular movies (by `popularity` score) in the **Action** genre. Sort them descending and display their title, vote average, and popularity.

::: {.callout-tip collapse="true" title="Answer Q3"}
```r
popular_df_genres |>
  filter(genre_name == "Action") |>
  arrange(desc(popularity)) |>
  select(title, vote_average, popularity) |>
  head(5)
```
The `popularity` field on TMDB is a proprietary score based on views, watchlists, and ratings — not the same as `vote_average`, which reflects quality. A movie can be very popular but not highly rated, and vice versa.
:::
#### ———————
