---
title: "Session 3: Weather API — First Real Requests"
subtitle: "Part B — Friday, August 7, 10:00–10:55 AM · Columbus I & II"
author: "Immanuel Williams"
date: "August 7, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    code-tools: true
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## Welcome Back — Part B

Today we use the free [Open-Meteo API](https://open-meteo.com/) to request and format weather data. Run the prepared code, inspect it, and complete the blanks.

::: {.callout-important}
## 🎯 The One Goal — Today We Do It

> **Make a request to an API, then format the response into something we can analyze.**

By the end, you will do this with live weather data.
:::


---


## Part 1: Investigating the URL

### Browser-request demonstration

**Stop. This is the moment the last two sessions have been building toward.** Here is a complete, real API request — and you can already read every piece of it from Session 2:

```
https://api.open-meteo.com/v1/forecast?latitude=35.28&longitude=-120.66&current=temperature_2m&temperature_unit=fahrenheit
```

Before you run anything, **inspect the URL**: identify the endpoint, location, requested variable, and units. Then select **Copy URL**, paste it into a browser, and press Enter. The result is raw JSON.

::: {.callout-important}
## Investigate Before You Run

Predict what the server will return from the URL alone. Only then press **Run** in R. Keep the **R console visible** so everyone can see the request, status, and response.
:::

``` json
{
  "latitude": 35.28,
  "longitude": -120.66,
  "timezone": "GMT",
  "current_units": { "time": "iso8601", "temperature_2m": "°F" },
  "current": {
    "time": "2026-08-06T16:00",
    "temperature_2m": 71.4
  }
}
```

### Read the request

- `https://api.open-meteo.com/v1/forecast` — the **endpoint** (the stable front door)
- `?` — "here come my details"
- `latitude=35.28&longitude=-120.66` — *where* (San Luis Obispo)
- `current=temperature_2m` — *what data* (`temperature_2m`)
- `temperature_unit=fahrenheit` — *how* (in °F)

The browser sends **GET** and returns JSON. Find `current`, then `temperature_2m`.

Change the coordinates to Boston (`42.36`, `-71.06`) and reload. Ask: **How does its weather compare with San Luis Obispo?** This is the question we will answer across cities and dates.

### About Open-Meteo

- No sign-up or API key.
- Endpoint: `https://api.open-meteo.com/v1/forecast`
- The query string specifies location, measurements, and units.

::: {.callout-note}
## Website or API?

- `open-meteo.com` serves documentation for people.
- `api.open-meteo.com` serves data for programs.
:::

::: {.callout-note}
## What About API Keys, Clients & Secrets?

Other APIs may require credentials. Open-Meteo does not.
:::

::: {.callout-important}
## 🎯 The Goal — Half Met Already

The browser completed one request. R will repeat the same process and format the response for analysis.
:::


#### ———————
### Q1. {.unnumbered}

**(Fill in the Blank):** The stable part of the URL — `https://api.open-meteo.com/v1/forecast` — that never changes from one request to the next is called the \_\_\_\_\_\_\_\_.

::: {.callout-tip collapse="true" title="Answer Q1"}
The **endpoint**. The changing part (after the `?`) is the query string.
:::
#### ———————


#### ———————
### Q2. {.unnumbered}

**(Multiple Choice):** Which statement about the Open-Meteo API is true?

a) It requires a paid subscription.
b) It requires an API key stored in `.Renviron`.
c) It is free and requires no API key at all.
d) It only works inside Positron.
e) It only returns data for cities in Europe.

::: {.callout-tip collapse="true" title="Answer Q2"}
**c)** Open-Meteo is free and open — no key, no registration. (Many other APIs *do* require a key; that's the "ID card" concept from the callout above.)
:::
#### ———————


#### ———————
### Q3. {.unnumbered}

**(Fill in the Blank):** When you paste the request URL into a browser, what comes back is not a web page — it is raw \_\_\_\_\_\_\_\_.

::: {.callout-tip collapse="true" title="Answer Q3"}
**JSON** — the same body/content structure we studied in Session 1.
:::
#### ———————


#### ———————
### Q4. {.unnumbered}

**(Multiple Choice):** Typing a URL into the browser's address bar and pressing Enter performs which kind of HTTP request?

a) POST
b) DELETE
c) PUT
d) PATCH
e) GET

::: {.callout-tip collapse="true" title="Answer Q4"}
**e)** The address bar always makes a **GET** request — you are *reading* data, exactly like our R scripts will.
:::
#### ———————


---


## Part 2: The ETV Cycle + Extract in R

### Look First — The Map for the Rest of the Workshop

<!-- TODO(Immanuel): link/embed your ETV Cycle 4 notes here -->

Everything follows **ETV**:

<details>
<summary>Show the ETV reference table</summary>

| Step | Name | What it means | In code |
|------|------|----------------|---------|
| **E** | **Extract** | Make the request, receive the JSON response | `request()` → `req_url_query()` → `req_perform()` |
| **T** | **Transform** | Format the response into a data frame | `resp_body_json()` → `tibble()` |
| **V** | **Visualize** | Turn the data frame into a plot | `ggplot()` |

</details>

### Now in Words

Today: **Extract** and **Transform**. Session 4 adds **Visualize**.

### Load Libraries

Load the tools for each ETV step:

```{r}
#| output: false
library(httr2)       # E: makes web requests
library(tibble)      # T: tidyverse data frames
library(dplyr)       # T: data manipulation
library(purrr)       # T: map() — apply a recipe to many inputs
library(ggplot2)     # V: visualizations
```

### Function reference

<details>
<summary>Show the ETV function table</summary>

| Library | Key functions | ETV step | What it actually does |
|---------|---------------|----------|------------------------|
| `httr2` | `request()`, `req_url_query()`, `req_perform()` | **E** | `request()` starts from the endpoint, `req_url_query()` builds the query string (the part after `?`), and `req_perform()` actually sends it — together they do what your browser's address bar did in Part 1 |
| `httr2` | `resp_status()`, `resp_body_json()` | **E → T** (the handoff) | `resp_status()` checks for `200`, and `resp_body_json()` unwraps the JSON body into an R named list — the bridge from *response* to *data* |
| `tibble` | `tibble()`, `tribble()` | **T** | Builds the rectangular data frame — the "something we can analyze" from the goal sentence |
| `dplyr` | `bind_rows()` | **T** | Stacks many small tibbles into one big one (we'll need this the moment we ask about more than one city) |
| `purrr` | `map()`, `pmap()` | **T** (the multiplier) | Applies the same E+T recipe to many inputs — this is the "automation" that makes R better than the browser |
| `ggplot2` | `ggplot()` | **V** | Turns the tibble into a plot — Session 4's job |

</details>

::: {.callout-tip}
## 🎯 Goal Check
**E** requests data, **T** formats it, and **V** displays it.
:::

### Extract — The Browser Request, Now in R

Build the same browser request one step at a time. Run each chunk and inspect `req` before continuing.

```{r}
# Step 1: Start with the endpoint
req <- request("https://api.open-meteo.com/v1/forecast")
req
```

```{r}
# Step 2: State the HTTP method explicitly
req <- req |>
  req_method("GET")
req
```

```{r}
# Step 3: Add the query string, including temperature_2m
req <- req |>
  req_url_query(
    latitude = 35.28,
    longitude = -120.66,
    current = "temperature_2m",
    temperature_unit = "fahrenheit"
  )
req
```

```{r}
# Step 4: Send the GET request
response <- req_perform(req)
```

```{r}
# Step 5: Check for success
resp_status(response)
```

The status should be `200` — success, just like Session 1 taught us to expect.

::: {.callout-note}
## `req_url_query()` Writes the URL for You
Each argument becomes `key=value`. `httr2` adds `?`, `&`, and URL encoding.
:::

```{r}
# Step 6: Parse JSON into a named list
result <- resp_body_json(response)

# Step 7: Inspect the list
str(result, max.level = 2)
```


#### ———————
### Q5. {.unnumbered}

**(Fill in the Blank):** After `req_perform()`, we call `resp_status(response)` and hope to see the number \_\_\_\_\_\_\_\_.

::: {.callout-tip collapse="true" title="Answer Q5"}
**200** — the "OK" success code. Anything in the 4xx range would mean *we* made a mistake in the request; 5xx would mean the server had a problem.
:::
#### ———————


#### ———————
### Q6. {.unnumbered}

**(Multiple Choice):** What does `req_url_query()` do?

a) It sends the request to the server.
b) It parses the JSON response.
c) It checks the status code.
d) It adds `key=value` pairs to the URL's query string, handling `?`, `&`, and encoding for you.
e) It draws a plot of the response.

::: {.callout-tip collapse="true" title="Answer Q6"}
**d)** `req_url_query()` *builds* the query string onto the endpoint. Sending is `req_perform()`, parsing is `resp_body_json()`.
:::
#### ———————


---


## Part 3: Transform the Response

::: {.panel-tabset}

### Main lesson: one city

The parsed `result` is a **named list** (Session 1!). The piece we care about is `result$current`:

```{r}
# T is for Transform ----------------------------------------

result$current

# Pull the values we want into a tibble
current_df <- tibble(
  city   = "San Luis Obispo",
  time   = result$current$time,
  temp_f = result$current$temperature_2m
)

current_df
```

The request is now a one-row data frame that we can compare later.

### Optional: `map()` and `pmap()`

Use this optional tab to see how one recipe can run for several cities.

**This block is pre-built.** Don't type it — **run it, watch it work, then read it** with the note below. (In your workshop file, a few blanks mark the values you choose; everything structural is already written.)

```{r}
# A small table of cities and their coordinates
cities <- tribble(
  ~city,             ~latitude, ~longitude,
  "San Luis Obispo",     35.28,    -120.66,
  "Boston",              42.36,     -71.06,
  "Ames",                42.03,     -93.62
)

# The recipe: for one row (one city), do E and T, return a one-row tibble
current_df <- cities |>
  pmap(\(city, latitude, longitude) {

    # E: extract
    result <- request("https://api.open-meteo.com/v1/forecast") |>
      req_url_query(
        latitude = latitude,
        longitude = longitude,
        current = "temperature_2m",
        temperature_unit = "fahrenheit"
      ) |>
      req_perform() |>
      resp_body_json()

    # T: transform
    tibble(
      city   = city,
      time   = result$current$time,
      temp_f = result$current$temperature_2m
    )
  }) |>
  bind_rows()

current_df
```

::: {.callout-note}
## Reading the `map` Pattern

- `\(city, latitude, longitude) { ... }` is an **anonymous function** — a recipe with no name. The `\( )` lists the ingredients; the `{ }` holds the steps.
- `pmap()` runs the recipe once **per row** of `cities`.
- Each run returns a one-row tibble; `bind_rows()` stacks them into one data frame.

Same ETV steps as before — just repeated for us, without copy-paste and without defining named functions. (Writing this recipe as a named function yourself is a great [take-home exercise](take_home_exercises.qmd).)
:::


#### ———————
#### Q7. {.unnumbered}

**(Multiple Choice):** In the code above, what does `\(city, latitude, longitude)` mean?

a) It creates three new columns in the tibble.
b) It is an anonymous function — a nameless recipe whose inputs are one row's city, latitude, and longitude.
c) It deletes the variables city, latitude, and longitude.
d) It is a comment that R ignores.
e) It is a file path.

::: {.callout-tip collapse="true" title="Answer Q7"}
**b)** `\( )` is R's shorthand for `function( )`. It defines a small nameless function on the spot, which `pmap()` then applies to every row.
:::
#### ———————


#### ———————
#### Q8. {.unnumbered}

**(Fill in the Blank):** After `pmap()` returns a list of one-row tibbles, we stack them into a single data frame using \_\_\_\_\_\_\_\_.

::: {.callout-tip collapse="true" title="Answer Q8"}
`bind_rows()` — it binds the rows of each small tibble into one combined tibble.
:::
#### ———————

:::

---


## Part 4: Geocoding + Your Turn

The **geocoding** endpoint converts a city name into coordinates.

### Look First — A Different Door

::: {.callout-important}
## ⚠️ Look Closely — This Is a *Different Endpoint*

Compare the endpoints:

- Forecast: `https://api.open-meteo.com/v1/forecast`
- Geocoding: `https://geocoding-api.open-meteo.com/v1/search`

Different endpoints accept different parameters and return different JSON structures.
:::

Inspect this Boston request, then open it in a browser:

```
https://geocoding-api.open-meteo.com/v1/search?name=Boston&count=1
```

### Now in Words

The first result contains Boston's coordinates. Make that same request in R:

```{r}
# Build the geocoding request for one city
geo_req <- request("https://geocoding-api.open-meteo.com/v1/search") |>
  req_method("GET") |>
  req_url_query(name = "Boston", count = 1)

# Send the request and parse its JSON body
geo_result <- geo_req |>
  req_perform() |>
  resp_body_json()

# Select the first matching place
hit <- geo_result$results[[1]]

# Save the coordinates in a one-row table
coords_df <- tibble(
  city = "Boston",
  latitude = round(hit$latitude, 2),
  longitude = round(hit$longitude, 2)
)

# Display the result
coords_df
```

We can now find coordinates for any city.

::: {.callout-tip}
## 🎯 Goal Check
Different endpoint, same ETV pattern.
:::


#### ———————
### Q9. {.unnumbered}

**(Fill in the Blank):** The geocoding endpoint converts a city name into \_\_\_\_\_\_\_\_ and \_\_\_\_\_\_\_\_.

::: {.callout-tip collapse="true" title="Answer Q9"}
**Latitude** and **longitude** — the coordinates the forecast and historical endpoints require.
:::
#### ———————


### Your Turn — Build Your Own Request

Make a current weather request for **your hometown** (or any city you choose). The code is scaffolded — fill in the city, complete endpoint, and response elements.

### Activity: Fill in the Blanks

::: {.callout-note}
## Inside the “black box”

The pipeline hides several small actions: build a GET URL, send it, receive JSON, select values from the nested list, and place those values in a tibble. Before running the code, point to the line responsible for each action.
:::

```{r}
# Step 1: Find your city's coordinates with the geocoding endpoint
my_coords <- request("https://geocoding-api.open-meteo.com/v1/search") |>
  req_method("GET") |>
  req_url_query(name = "________", count = 1) |>
  req_perform() |>
  resp_body_json()

# Pull latitude and longitude from the first match
my_lat <- my_coords$results[[1]]$________
my_lon <- my_coords$results[[1]]$________

# Step 2: Request the current weather (E)
my_result <- request("________") |>
  req_method("GET") |>
  req_url_query(
    latitude = ________,
    longitude = ________,
    current = "temperature_2m",
    temperature_unit = "fahrenheit"
  ) |>
  req_perform() |>
  resp_body_json()

# Step 3: Format the response into a tibble (T)
my_df <- tibble(
  city   = "________",
  time   = my_result$current$________,
  temp_f = my_result$current$________
)

my_df
```

::: {.callout-tip collapse="true" title="Solution"}
```{r}
my_coords <- request("https://geocoding-api.open-meteo.com/v1/search") |>
  # Send a GET request for Chicago's first geocoding match
  req_method("GET") |>
  req_url_query(name = "Chicago", count = 1) |>
  req_perform() |>
  resp_body_json()

# Pull coordinates from the first result
my_lat <- my_coords$results[[1]]$latitude
my_lon <- my_coords$results[[1]]$longitude

# Request Chicago's current temperature
my_result <- request("https://api.open-meteo.com/v1/forecast") |>
  req_method("GET") |>
  req_url_query(
    latitude = my_lat,
    longitude = my_lon,
    current = "temperature_2m",
    temperature_unit = "fahrenheit"
  ) |>
  req_perform() |>
  resp_body_json()

# Convert the nested response values to a tibble
my_df <- tibble(
  city   = "Chicago",
  time   = my_result$current$time,
  temp_f = my_result$current$temperature_2m
)

# Display the finished row
my_df
```
:::


---


## Session 3 Recap

1. Inspect a GET URL and its JSON response.
2. Build and send the same request in R.
3. Transform nested values into a tibble.
4. Geocode a city and request its current temperature.

Optional: use `map()` or `pmap()` to repeat a request.

**Up next in Session 4 — Historical Weather, Visualization & Wrap-Up:** the **V** in ETV — historical weather data for Boston vs. San Luis Obispo, and your own visualizations!


---


## During the Break {.unnumbered}

Before Session 4 starts:

- **Questions?** Come up and ask — Meg and I will be here for the whole break.
- **Did you finish?** Check yourself:
    - [ ] I pasted the forecast URL into my browser and saw JSON
    - [ ] I ran the R request and got a `200` status
    - [ ] I built `my_df` for my own city in Part 4
- If you finished everything, try adding a fourth city to the `cities` table in Part 3 and re-running the `pmap()` block.
