---
title: "Session 4: Historical Weather, Visualization & Wrap-Up"
subtitle: "Day 2 — 10:00 AM to 11:00 AM"
author: "Immanuel Williams"
date: "August 6, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    code-tools: true
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## Recap from Session 3

In Session 3, we:

- Made our first real requests to the free **Open-Meteo** API (no API key!)
- Saw the request in the **browser first**, then automated it in R
- Learned the **ETV Cycle**: Extract → Transform → Visualize
- Used `map()` + anonymous functions `\( )` to repeat the recipe for many cities
- Used the **geocoding** endpoint to turn city names into coordinates

Today we complete the cycle: **historical** weather data and the **V** — visualization.

::: {.callout-important}
## 🎯 The One Goal — Final Step

> **Make a request to an API, then format the response into something we can analyze.**

Today "analyze" becomes concrete: real plots comparing real cities.
:::

### Load Libraries

```{r}
#| output: false
library(httr2)       # E: makes web requests
library(tibble)      # T: tidyverse data frames
library(dplyr)       # T: data manipulation
library(purrr)       # T: map() — apply a recipe to many inputs
library(ggplot2)     # V: visualizations
```


---


## Part 1: Extract — Historical Weather Data

Open-Meteo has a second endpoint for **past** weather — the archive endpoint:

```
https://archive-api.open-meteo.com/v1/archive
```

Browser first, as always. Paste this — it asks for San Luis Obispo's daily high and low temperatures for one week in July 2025:

```
https://archive-api.open-meteo.com/v1/archive?latitude=35.28&longitude=-120.66&start_date=2025-07-01&end_date=2025-07-07&daily=temperature_2m_max,temperature_2m_min&temperature_unit=fahrenheit&timezone=auto
```

Two new query parameters do the date work: `start_date` and `end_date`. And here's the good news — **one request returns the entire date range**. No loops needed.

Now in R:

```{r}
# E: extract -------------------------------------------------
result <- request("https://archive-api.open-meteo.com/v1/archive") |>
  req_url_query(
    latitude = 35.28,
    longitude = -120.66,
    start_date = "2025-07-01",
    end_date = "2025-07-07",
    daily = "temperature_2m_max,temperature_2m_min",
    temperature_unit = "fahrenheit",
    timezone = "auto"
  ) |>
  req_perform() |>
  resp_body_json()

# Peek at the daily piece
str(result$daily, max.level = 1)
```

The `daily` element is a named list of three parallel lists: `time`, `temperature_2m_max`, and `temperature_2m_min` — one entry per day.

```{r}
# T: transform ------------------------------------------------
slo_df <- tibble(
  city     = "San Luis Obispo",
  date     = as.Date(unlist(result$daily$time)),
  temp_max = unlist(result$daily$temperature_2m_max),
  temp_min = unlist(result$daily$temperature_2m_min)
)

slo_df
```

::: {.callout-note}
## Why `unlist()`?
`resp_body_json()` parses each JSON array into an R **list**. `unlist()` flattens that list into a plain **vector**, which is what a tibble column needs. List in, vector out — that's all it does.
:::


#### ———————
### Q1. {.unnumbered}

**(Fill in the Blank):** To ask the archive endpoint for a different range of days, we change the \_\_\_\_\_\_\_\_ and \_\_\_\_\_\_\_\_ query parameters.

::: {.callout-tip collapse="true" title="Answer Q1"}
`start_date` and `end_date`. Everything else in the request can stay exactly the same.
:::
#### ———————


#### ———————
### Q2. {.unnumbered}

**(Multiple Choice):** What does `unlist()` do in the Transform step?

a) It flattens a list into a plain vector so it can be a tibble column.
b) It sends the request to the server.
c) It removes missing values from the data.
d) It converts Fahrenheit to Celsius.
e) It sorts the dates in order.

::: {.callout-tip collapse="true" title="Answer Q2"}
**a)** JSON arrays arrive in R as lists; `unlist()` turns each one into a vector — exactly what tibble columns are made of.
:::
#### ———————


---


## Part 2: Visualize — Completing the ETV Cycle

We have a tidy data frame. Time for **V**:

```{r}
# V: visualize -------------------------------------------------
ggplot(slo_df, aes(x = date)) +
  geom_line(aes(y = temp_max, color = "High"), linewidth = 1.2) +
  geom_line(aes(y = temp_min, color = "Low"), linewidth = 1.2) +
  scale_color_manual(values = c("High" = "red", "Low" = "blue")) +
  labs(
    title = "Daily High and Low Temperatures in San Luis Obispo",
    subtitle = "July 1–7, 2025 (Open-Meteo archive)",
    x = "Date",
    y = "Temperature (°F)",
    color = "Temperature"
  ) +
  theme_minimal(base_size = 14)
```

🎉 **That's the entire workshop goal, end to end:** we made a request, formatted the response, and analyzed it with a plot. Everything from here is variations on this cycle.


#### ———————
### Q3. {.unnumbered}

**(Fill in the Blank):** In the ETV cycle, `ggplot()` is the tool for the \_\_\_\_\_\_\_\_ step.

::: {.callout-tip collapse="true" title="Answer Q3"}
The **Visualize** step — the V. (Extract was `request()`/`req_perform()`, Transform was `resp_body_json()`/`tibble()`.)
:::
#### ———————


---


## Part 3: Boston vs. San Luis Obispo — Two Simple Comparisons

Let's use the `map()` pattern from Session 3 to compare two cities. The recipe below fetches one city's daily highs for any date range:

```{r}
cities <- tribble(
  ~city,             ~latitude, ~longitude,
  "San Luis Obispo",     35.28,    -120.66,
  "Boston",              42.36,     -71.06
)
```

### Comparison 1: Summer

```{r}
# E + T for both cities: July 2025
summer_df <- cities |>
  pmap(\(city, latitude, longitude) {

    result <- request("https://archive-api.open-meteo.com/v1/archive") |>
      req_url_query(
        latitude = latitude,
        longitude = longitude,
        start_date = "2025-07-01",
        end_date = "2025-07-31",
        daily = "temperature_2m_max",
        temperature_unit = "fahrenheit",
        timezone = "auto"
      ) |>
      req_perform() |>
      resp_body_json()

    tibble(
      city     = city,
      date     = as.Date(unlist(result$daily$time)),
      temp_max = unlist(result$daily$temperature_2m_max)
    )
  }) |>
  bind_rows()

# V: one line per city
ggplot(summer_df, aes(x = date, y = temp_max, color = city)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Summer Daily Highs: Boston vs. San Luis Obispo",
    subtitle = "July 2025",
    x = "Date",
    y = "Daily High (°F)",
    color = "City"
  ) +
  theme_minimal(base_size = 14)
```


#### ———————
### Q4 — Visual Check (Summer). {.unnumbered}

**(Multiple Choice):** In the summer plot, what distinguishes Boston's line from San Luis Obispo's line?

a) The line type (solid vs. dashed)
b) The x-axis position
c) Separate panels for each city
d) The point shape
e) The line color, mapped with `aes(color = city)`

::: {.callout-tip collapse="true" title="Answer Q4"}
**e)** We mapped the `city` column to the `color` aesthetic — `aes(color = city)` — so ggplot automatically draws one colored line per city and builds the legend.
:::
#### ———————


### Comparison 2: Winter

Same recipe — the **only change is the two dates**:

```{r}
# E + T for both cities: January 2026
winter_df <- cities |>
  pmap(\(city, latitude, longitude) {

    result <- request("https://archive-api.open-meteo.com/v1/archive") |>
      req_url_query(
        latitude = latitude,
        longitude = longitude,
        start_date = "2026-01-01",
        end_date = "2026-01-31",
        daily = "temperature_2m_max",
        temperature_unit = "fahrenheit",
        timezone = "auto"
      ) |>
      req_perform() |>
      resp_body_json()

    tibble(
      city     = city,
      date     = as.Date(unlist(result$daily$time)),
      temp_max = unlist(result$daily$temperature_2m_max)
    )
  }) |>
  bind_rows()

# V
ggplot(winter_df, aes(x = date, y = temp_max, color = city)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Winter Daily Highs: Boston vs. San Luis Obispo",
    subtitle = "January 2026",
    x = "Date",
    y = "Daily High (°F)",
    color = "City"
  ) +
  theme_minimal(base_size = 14)
```


#### ———————
### Q5 — Visual Check (Winter). {.unnumbered}

**(Multiple Choice):** Comparing the summer and winter code blocks, what changed between them?

a) The endpoint changed to a winter-specific API.
b) An API key was added for winter data.
c) Only the `start_date` and `end_date` values changed.
d) The cities table had to be rebuilt.
e) `pmap()` was replaced by a `for` loop.

::: {.callout-tip collapse="true" title="Answer Q5"}
**c)** Two date strings — that's the entire difference. Same endpoint, same recipe, same transform, same plot code. This is the power of the ETV pattern: new question, two small edits.
:::
#### ———————


#### ———————
### Q6. {.unnumbered}

**(Fill in the Blank):** In both comparison plots, we expect \_\_\_\_\_\_\_\_ to show much colder daily highs in January, while San Luis Obispo stays mild year-round.

::: {.callout-tip collapse="true" title="Answer Q6"}
**Boston.** Coastal California has a mild Mediterranean climate; Boston winters regularly stay below freezing — the winter plot makes the gap dramatic, often 30–40°F.
:::
#### ———————


---


## Part 4: Explore — Build Your Own Query

Now **you** design the question. Pick **one option from each column**, then fill in the template below and make your plot.

| 🏙 City (lat, lon) | 📊 Daily variable | 📅 Date range |
|---|---|---|
| Chicago (41.88, -87.63) | `temperature_2m_max` — daily high | Summer: `2025-06-01` → `2025-08-31` |
| Miami (25.76, -80.19) | `temperature_2m_min` — daily low | Winter: `2025-12-01` → `2026-02-28` |
| Denver (39.74, -104.98) | `precipitation_sum` — total precipitation | Spring: `2026-03-01` → `2026-05-31` |
| Seattle (47.61, -122.33) | `windspeed_10m_max` — peak wind | Last 3 months: `2026-04-01` → `2026-06-30` |
| Ames (42.03, -93.62) | `snowfall_sum` — snowfall | One month of your choice |
| *Your hometown (geocode it!)* | | |

### Activity: Fill in the Blanks

```{r}
# Your choices
my_result <- request("https://archive-api.open-meteo.com/v1/archive") |>
  req_url_query(
    latitude = ________,
    longitude = ________,
    start_date = "________",
    end_date = "________",
    daily = "________",
    temperature_unit = "fahrenheit",
    timezone = "auto"
  ) |>
  req_perform() |>
  resp_body_json()

my_df <- tibble(
  date  = as.Date(unlist(my_result$daily$time)),
  value = unlist(my_result$daily$________)
)

ggplot(my_df, aes(x = date, y = value)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  labs(
    title = "________",
    x = "Date",
    y = "________"
  ) +
  theme_minimal(base_size = 14)
```

::: {.callout-tip collapse="true" title="Example Solution — Seattle's Spring Winds"}
```{r}
my_result <- request("https://archive-api.open-meteo.com/v1/archive") |>
  req_url_query(
    latitude = 47.61,
    longitude = -122.33,
    start_date = "2026-03-01",
    end_date = "2026-05-31",
    daily = "windspeed_10m_max",
    temperature_unit = "fahrenheit",
    timezone = "auto"
  ) |>
  req_perform() |>
  resp_body_json()

my_df <- tibble(
  date  = as.Date(unlist(my_result$daily$time)),
  value = unlist(my_result$daily$windspeed_10m_max)
)

ggplot(my_df, aes(x = date, y = value)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  labs(
    title = "Peak Daily Wind Speed in Seattle, Spring 2026",
    x = "Date",
    y = "Max Wind Speed (mph)"
  ) +
  theme_minimal(base_size = 14)
```
:::

**Bonus:** Compare *two* cities on your chosen variable — copy the `pmap()` recipe from Part 3 and swap in your variable.


#### ———————
### Q7. {.unnumbered}

**(Discussion):** You just built a complete data pipeline: API request → JSON parsing → data wrangling → visualization. Think about the courses you teach. Where could API data naturally integrate into your curriculum? What topics could benefit from live, real-world data?

::: {.callout-tip collapse="true" title="Example Answers"}
Possible integration points:

- **Intro Stats:** Use weather data for descriptive statistics (mean temperature, variability across cities)
- **Regression:** Model temperature as a function of latitude or time
- **Data Wrangling:** Parse nested JSON, clean and reshape API responses
- **Probability:** Explore weather prediction accuracy (forecast vs. actual)
- **Capstone Projects:** Let students choose an API and build their own data pipeline
:::
#### ———————


---


## Part 5: Reflection & Pitfalls

### Common Pitfalls When Working with APIs

1. **Rate Limits:** Most APIs limit how many requests you can make per minute/hour (Open-Meteo's free tier allows ~10,000 calls/day — generous, but not infinite). Design your code to respect limits.

2. **API Keys (elsewhere):** Most commercial APIs require keys. When you use one, never write the key directly in your script — store it in an environment file. The Practice Labs on this site walk through this.

3. **Data Structure Changes:** API providers can change their response format without warning. Always inspect the structure of new responses (`str()`) before building pipelines around them.

4. **URL Encoding:** Special characters (spaces, ampersands, accents) must be encoded in URLs. `req_url_query()` handles this for you automatically — one more reason to like it.

5. **Nested JSON Complexity:** Real-world JSON can be deeply nested. Practice with `str()` and `names()` to navigate the structure before trying to flatten it.


### Reflection Questions for Your Practice

- Where could API data naturally integrate into your curriculum?
- What are the pitfalls (rate limits, authentication) that students need to know?
- How does working with live, messy data differ from working with pre-cleaned datasets?
- What ethical considerations arise when collecting data from APIs?


---


## Workshop Summary

| Session | Day | Time | Key Topics |
|---------|-----|------|------------|
| 1 | Day 1 | 8–9 AM | JSON fundamentals, request/response, status codes |
| 2 | Day 1 | 9–10 AM | CRUD, GET/POST, URL anatomy |
| 3 | Day 2 | 9–10 AM | Open-Meteo requests in browser & R, ETV cycle, geocoding |
| 4 | Day 2 | 10–11 AM | Historical weather, city comparisons, your own query |

**Key Takeaways:**

1. APIs provide **programmatic access** to live data
2. Requests are **URLs** — responses are **JSON**
3. Status codes help you **diagnose problems**
4. The **ETV cycle** (Extract → Transform → Visualize) is the same for every API
5. `map()` + anonymous functions repeat the recipe **without copy-paste**
6. Live data creates **authentic learning experiences**

Thank you for participating! Take these skills and bring real-world data into your classrooms.


---


## During the Break — and After the Workshop {.unnumbered}

- **Questions?** Come up and ask — Meg and I will stick around.
- **Did you finish?** Check yourself:
    - [ ] I ran the historical request and plotted San Luis Obispo's July temperatures
    - [ ] I made both Boston vs. SLO comparison plots
    - [ ] I built and plotted **my own query** in Part 4
- **Resources for when this class is over:**
    - **Practice Labs on this site** — five guided labs (NASA, Cat API, OpenWeather, TMDB, GitHub), including APIs that *do* require keys
    - **Session 2's "Optional — After the Workshop" section** — dummy-server practice with GET/POST and error handling
    - [Open-Meteo documentation](https://open-meteo.com/en/docs) — every variable and endpoint we used, and many more
    - [httr2 package site](https://httr2.r-lib.org/) — the full request/response toolkit
    - [R for Data Science (2e), Iteration chapter](https://r4ds.hadley.nz/iteration.html) — more on `map()` and friends
    - [JSONPlaceholder](https://jsonplaceholder.typicode.com/) — the free dummy server, for classroom practice
