---
title: "Take-Home Exercises: Going Deeper"
subtitle: "Optional practice for after the workshop"
author: "Immanuel Williams"
date: "August 5, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    code-tools: true
execute:
  echo: true
  eval: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

::: {.callout-warning}
## Read This Later, Not During the Workshop

The three exercises below are **optional enrichment** — we do *not* cover them live. They preview the R code from Sessions 3–4 against a practice server and go deeper into error handling. Nothing in the live sessions depends on them; work through them at your own pace after the workshop.
:::

These exercises originally lived at the end of [Session 2 — The Anatomy of a Request](02_CRUD_HTTP_Methods_URL.qmd). They are also a good place to practice **building functions yourself** — during the workshop you mostly *used* pre-built code, and here you get to write some from scratch.


## Exercise A: Under the Hood — What Happens When You Make a Request

Let's connect everything together:

1. You **construct a URL** from a base URL (endpoint) + query string (parameters).
2. Your client sends a **GET request** using this URL to the API.
3. The API relays this request to the **server**.
4. The server processes the request, queries the database, and sends a **response**.
5. The response includes a **status code** and the requested data (formatted as **JSON**).
6. Your R code **parses** the JSON into a data structure you can work with.

::: {.callout-important}
## Common Misconception
We are **not** sending JSON when we make a GET request. We're sending a URL with parameters. **JSON is what comes back** to us in the response.
:::


## Exercise B: Making Requests to a Dummy Server in R

These blocks show real requests to a practice server. You can run them any time after the workshop.

```{r}
#| output: false
library(httr2)       # Makes web requests
library(tibble)      # Tidyverse data frames
library(dplyr)       # Data manipulation
library(glue)        # String construction
```

A simple GET request to a public test API:

```{r}
# Step 1: Define the URL
test_url <- "https://jsonplaceholder.typicode.com/users/1"

# Step 2: Create a request object
req <- request(test_url)

# Step 3: Perform the request
response <- req_perform(req)

# Step 4: Check the status code
resp_status(response)

# Step 5: Check the content type
resp_content_type(response)
```

```{r}
# Step 6: Parse the JSON response
result <- resp_body_json(response)

# Step 7: Inspect the structure
str(result)
```

```{r}
# Step 8: Convert to a data frame
user_df <- as.data.frame(result)
print(user_df)
```

A GET request with query parameters:

```{r}
# Get posts by a specific user
posts_url <- "https://jsonplaceholder.typicode.com/posts"

posts_url_query <- glue("{posts_url}?userId=1")

req <- request(posts_url_query)
response <- req_perform(req)

if (resp_status(response) == 200) {
  posts <- resp_body_json(response)
  cat("Number of posts retrieved:", length(posts), "\n")
}
```

And a POST request — **sending** data to the server:

```{r}
# Step 1: Define the endpoint
post_url <- "https://jsonplaceholder.typicode.com/posts"

# Step 2: Create the request with a body
req <- request(post_url) |>
  req_method("POST") |>
  req_headers("Content-Type" = "application/json") |>
  req_body_json(list(
    title = "My API Workshop Post",
    body = "Learning about APIs is fun!",
    userId = 1
  ))

# Step 3: Perform the request
response <- req_perform(req)

# Step 4: Check the status code (201 = Created)
cat("Status:", resp_status(response), "\n")

# Step 5: See the response
result <- resp_body_json(response)
print(result)
```

::: {.callout-note}
## Status Code 201
Notice we got a `201 Created` instead of `200 OK`. This is the standard success code for POST requests — it means the server successfully created the new resource.
:::


## Exercise C: Handling Errors with Status Codes

```{r}
# Try requesting a resource that doesn't exist
bad_url <- "https://jsonplaceholder.typicode.com/users/999999"

req <- request(bad_url)
response <- req_perform(req)

# Check the status
cat("Status code:", resp_status(response), "\n")
```

```{r}
# Build a robust request handler
safe_get <- function(url) {
  req <- request(url)
  response <- req_perform(req)
  status <- resp_status(response)

  if (status == 200) {
    cat("Success! Status:", status, "\n")
    return(resp_body_json(response))
  } else if (status >= 400 & status < 500) {
    cat("Client error! Status:", status, "- Check your URL or parameters.\n")
    return(NULL)
  } else if (status >= 500) {
    cat("Server error! Status:", status, "- The server has an issue.\n")
    return(NULL)
  }
}

# Test it
result <- safe_get("https://jsonplaceholder.typicode.com/users/1")
```

#### ———————
### Bonus Q. {.unnumbered}

**(Open-Ended):** In the GET request to the dummy server, we sent a URL and received JSON. In the POST request, we sent JSON *in the request body*. How does this change what we learned about GET vs. POST?

::: {.callout-tip collapse="true" title="Answer — Bonus Q"}
This reinforces the key difference: **GET** requests encode parameters in the URL and receive JSON back. **POST** requests send data in the request body (often as JSON) and receive a confirmation (also JSON) back. So POST actually involves JSON going *both directions*, while GET only involves JSON coming back.
:::
#### ———————


## More Practice

- **Practice Labs on this site** — five guided labs (NASA, Cat API, OpenWeather, TMDB, GitHub), including APIs that *do* require keys
- Rebuild the workshop's pre-built functions from scratch: write your own `geocode()` and `prev_weather()` wrappers around the Session 3–4 pipelines
- [Open-Meteo documentation](https://open-meteo.com/en/docs) · [httr2 package site](https://httr2.r-lib.org/) · [JSONPlaceholder](https://jsonplaceholder.typicode.com/)
